Mediator Subunit MED16 Collaborates with UBP1-TFCP2 to Control Transcriptional Activation or Repression via Promoter Positional Specificity

Version 1

Abstract

The Mediator complex is an evolutionarily conserved transcriptional coactivator with well-characterized structure and function, though the roles of its dissociable subunits remain incompletely defined. Here, we demonstrate that Mediator subunit MED16 dissociates from the core complex to form a subcomplex with transcription factors UBP1 and TFCP2, and this interaction modulates transcriptional activation or repression in a context-dependent manner. Using protein purification coupled with mass spectrometry, we identified MED16 as a binding partner of UBP1-TFCP2. Gene expression analyses showed that UBP1 interacts with MED16 to activate a subset of silenced genes involved in lung homeostasis, angiogenesis, and cell proliferation. Conversely, the MED16-UBP1 interaction suppresses HIV-1 transcription, thereby reinforcing viral latency. Mechanistically, MED16 and UBP1 cooperatively bind the HIV-1 transcriptional start site (TSS) to inhibit preinitiation complex assembly. Genomic-scale analyses further demonstrated that transcription is activated when the UBP1-TFCP2 binding motif is proximal to the TSS, but repressed when the motif overlaps the TSS. Collectively, our findings identify a novel MED16-UBP1 interaction, define its dual role in transcriptional regulation, and highlight the therapeutic potential of targeting this axis in HIV-1 infection.

Introduction

The canonical model for eukaryotic gene transcription by RNA polymerase II (Pol II) is that transcription factors recruit co-activators and RNA Pol II. Together, they are assembled into the pre-initiation complex (PIC) along with general transcription factors (GTFs) at a gene promoter (1-8). Among several kinds of co-activators reported, the multi-subunit Mediator complex has emerged as an essential co-activator that promotes PIC formation and activity.

The Mediator complex is composed of ∼30 subunits in mammals and structurally divided into four modules: Head, Middle, Tail, and Kinase. The scaffold subunit MED14 joins the Tail, Middle, and Head modules together to form the “core Mediator”. The tail module is composed of MED15, MED16, MED23, MED24, MED25, MED27, MED28, MED29, and MED30; and the main function of the tail module is to provide an interactional platform for transcription factors to recruit Mediator complex to gene promoters and enhancers for further regulation of gene expression (8-10). Among all tail subunits, MED16, MED23, MED24, and MED25 form a submodule that accounts for ∼70% molecular weight of the tail module and creates a large platform for interactions of transcription factors (11). Previous studies have found MED23 interacts with multiple transcriptional factors and controls diverse gene expression programs at multiple stages of transcription processes (12-15). It prompts us to question whether other subunits in this submodule also play similar roles in transcriptional regulation, especially MED16 that is structurally conserved from yeast to mammals and acts as a critical regulator in yeast and plants. Sin4 (yeast MED16) plays both positive and negative roles in the transcriptional regulation of many genes in yeast, and loss of Sin4 alters chromatin accessibility globally (16). In plant cells, MED16 controls endoreduplication and cell growth by repressing the expression of APC/C activator genes CCS52A1/A2 (17-21). MED16 also controls plant cold response, chemical and wound response, and iron homeostasis at transcriptional levels (17-21). However, the functions of MED16 in mammalian cells are poorly understood. Among a few reports, MED16 was reported as a key factor in NRF2-activted gene expression via its direct interaction with NRF2. Knockout of MED16 reduces NRF2-dependent anti-oxidative gene expression (21, 22). Additionally, MED16 was down-regulated in papillary thyroid cancer, leading to increased TGF-β signaling and radioiodine resistance (23). Though the exact molecular mechanisms by which MED16 regulates gene expression are poorly understood. Its roles in animal development and diseases are barely investigated.

UBP1 (as known as LBP-1) and TFCP2 (as known as LSF) belong to a subfamily of the TFCP2/Grainyhead transcription factors family. The amino acid sequences of TFCP2 and UBP1 share 88% identity, while TFCP2 and TFCP2L1 share 70% identity (24-26). UBP1 and TFCP2 bind to the same consensus sequence CNRG-N6-CNR(G/C) (27). They function as a transcription activator or repressor to regulate a number of viral and cellular genes, including mouse α-Globin, MMP-9, CYP11A1, IL-4, SV-40, and HIV-1 (28-32). UBP1 and TFCP2 are involved in various biological processes, including embryonic development, angiogenesis, tumorigenesis, and HIV-1 latency (28, 29, 31-38). There is a canonical UBP1 binding motif within the HIV-1 core promoter which overlaps with the transcription start site (TSS). Previous studies have reported that UBP1 inhibits the binding of TFIID to the TATA-box (32). On the other hand, UBP1 restricts HIV-1 transcription at the level of elongation (33). UBP1 also has been demonstrated to inhibit HIV-1 transcription by recruiting YY1 and HDAC1 (34). However, how UBP1 and TFCP2 regulate cellular gene expression, and what are the genome-wide target genes of UBP1 and TFCP2 still remain unclear.

In this study we identified a tripartite complex consisting of UBP1, TFCP2, and MED16 utilizing gel-filtration chromatography and immunoprecipitation-mass spectrometry (IP-MS). MED16 interacts with UBP1 to activate expression of an array of endogenous genes, including several pulmonary surface molecules. In addition, MED16 cooperates with UBP1 to inhibit HIV-1 transcription at the HIV-1transcription initiation site. At the genomic scale, we found out that UBP1 activates gene expression when the UBP1-TFCP2 binding motif is located near TSS, but inhibits gene expression by obstructing normal PIC formation when the UBP1-TFCP2 binding motif overlaps with TSS. Based on these results, we propose a dual functional model of MED16 in gene transcriptional control, and provided the underlying mechanisms of UBP1-repression on the HIV-1 genome, which implying a strategy for HIV-1 therapeutics.

Result

Identification of the dissociable MED16 from the Mediator complex

The Mediator complex consists of large intrinsically disordered regions (IDRs) and lacks predictable functional motifs (39, 40). The Mediator complex forms liquid-liquid phase separation condensates at super-enhancers. These IDR-mediated condensates can be disrupted by 1,6-hexanediol (1,6-HD) (41, 42), an aliphatic alcohol, via inhibition of weak hydrophobic protein-protein interactions (43-45).

To investigate whether 1,6-hexanediol affects the integrity of the Mediator complex, we utilized three antibodies against different Mediator subunits (CDK8, MED1 and MED12) to immunoprecipitate the Mediator complex from the 293T whole cell lysates, respectively. The immunoprecipitates were washed by washing buffers containing different concentrations (0%, 5%, 10%) of 1,6-hexanediol and blotted by antibodies against different Mediator components. Most of the detected subunits show modest decreases, whereas MED16 exhibits a pronounced reduction in response to increasing 1,6-hexanediol treatment (Fig 1A). To further confirm this observation, a similar experiment was performed in HeLa nuclear extract (NE) with 2,5-hexanediol as a control treatment. As an isomer of 1,6-hexanediol, 2,5-hexanediol does not interfere weak hydrophobic interaction due to the different positions of oxhydryls. Once again, the amount of immunoprecipitated MED16 was easily decreased by 1,6-hexanediol treatment but not by 2,5-hexanediol treatment (Fig 1B), suggesting that MED16 is an easily dissociable subunit.

MED16 dissociates from the Mediator complex and interacts with UBP1-TFCP2

(A) Co-IP experiment with antibodies against endogenous CDK8, MED1, and MED12 in 293T whole cell lysis. The resultant immunoprecipitates were washed by washing buffers containing 0%, 5%, and 10% of 1,6-hexanediol respectively. Input: 0.5% of the total lysate was loaded. (B) Co-IP experiment with antibodies against endogenous CDK8, MED1, and MED12 in HeLa nuclear extract. The resultant immunoprecipitates were washed by washing buffers containing 10% of 1,6-hexanediol or 10% 2,5-hexanediol. Input: 0.2% of the total HeLa nuclear extract was loaded. (C) Co-IP experiment with antibodies against endogenous MED1 in 293T whole cell lysis. The resultant immunoprecipitated was washed by washing buffers containing 150 mM, 300 mM, and 500 mM of NaCl respectively. Input: 0.2% of the total HeLa nuclear extract was loaded. (D) Gel filtration chromatography of HeLa nuclear extract. 500µl HeLa nuclear extract was applied to Superose 6 column then was run in Buffer D. Column fractions of 500µl were collected. The Void (void volumn) of Superose 6 is based on the volume of effluent required for the elution of blue dextran (molecular mass of ∼2000 kDa). The molecular weight of corresponding fractions was detected by protein standards. (E) Immunoblots of anti-MED16 immunoprecipitation in HeLa nuclear extract and gel filtration factions No.29 to No.31. The immunoprecipitated proteins were detected with indicated antibodies by western blotting. HeLa NE input: 0.2% of the total HeLa nuclear extract was loaded. (F) Immunoblots of anti-TFCP2 immunoprecipitation in HeLa nuclear extract and gel filtration factions No.29 to No.31. The immunoprecipitated proteins were detected with indicated antibodies by western blotting. HeLa NE input: 0.2% of the total HeLa nuclear extract was loaded.

We then immunoprecipitated the Mediator complex with MED1 antibody in high salt condition to test whether the binding of MED16 to Mediator complex is charge dependent. The amount of immunoprecipitated MED16 remained unaltered under 150 mM, 300 mM and 500 mM NaCl conditions (Fig 1C). These observations suggest that MED16 is integrated into the Mediator complex through weak hydrophobic interaction but not ionic interaction.

To directly examine if MED16 is dissociable from the Mediator complex, we applied HeLa NE to Superose 6 gel-filtration column and separated Mediator subunits within each fraction that were then detected by immunoblotting. Most Mediator subunits, including MED16, were co-eluted at the fraction at about 2,000 kDa, suggesting these fractions contain the integrated Mediator complex (Fig 1D). However, a significant amount of MED16 was eluted at the fraction between 443 kDa and 150 kDa, away from other core Mediator subunits, indicating MED16 within this fraction is not part of the core Mediator complex (Fig 1D). MED16 subunit may be able to dissociate from the Mediator complex because the interaction between Mediator and MED16 is an unstable hydrophobic interaction.

MED16 interacts with UBP1 and TFCP2 in vivo and in vitro

Because MED16 is about 95 kDa and eluted at the HeLa NE gel-filtration fractions between 443 kDa and 150 kDa, MED16 may form a higher molecular mass complex with other proteins. To identify the proteins associated with MED16, we performed immunoprecipitation-coupled mass spectrometry (IP-MS) analysis using antibodies against four Mediator subunits, including MED1, MED12, CDK8 and MED16, respectively. Forty-two proteins were identified as exclusively interacting with MED16 in IP-MS data. Among these, two dimeric transcription factors, UBP1 and TFCP2 (Fig S1), were found to be co-eluted with the dissociable MED16 in gel filtration chromatography (Fig 1D). This suggests that UBP1 and TFCP2 interact with the dissociated MED16 (Fig S1).

To determine whether UBP1 and TFCP2 form a complex with MED16, an antibody specific to MED16 was used to capture MED16 and its associated proteins in HeLa NE or HeLa NE gel-filtration fractions between 443 kDa and 150 kD. MED1, MED23, TFCP2 and UBP1 were co-immunoprecipitated with MED16 in HeLa NE, but only TFCP2 and UBP1 were co-immunoprecipitated with MED16 in NE fractions between 443KDa and 150KD (Fig 1E). Reciprocally, anti-TFCP2 antibody is also able to pull down MED16 in either HeLa NE or 443KDa-150KDa fractions (Fig 1F). YY1 and HDAC1 are known to complex with TFCP2 and UBP1 to suppress HIV-1 transcription (34, 46, 47). YY1 was pulled down in both samples, whereas only a small amount of HDAC1 was pulled down from HeLa NE but not from the 443KDa to 150KD fractions (Fig 1E).

Because TFCP2 and UBP1 can be co-immunoprecipitated with MED16 and other core Mediator subunits, we examined whether TFCP2 and UBP1 interact with the Mediator complex. Both the TFCP2 and UBP1 were co-imunoprecipitated with Mediator complex under the 2,5-hexanediol treatment, whereas their association with Mediator complex was abolished under treatment with 1,6-hexanediol (Fig 1B). These results suggest that TFCP2 and UBP1 are complexed with MED16 in vivo and dissociate from the Mediator complex, possibly along with the dissociation of MED16.

To extend the analysis of the MED16 subcomplex, a domain-mapping experiment with deletions of multiple structural domains of MED16 was carried out. The N-terminal of MED16 contains nine WD40-repeat (WDR) domains that form a seven-bladed β-propeller structure (Fig 2A) (11, 48). This structure appears to be crucial for MED16’s integration into the Mediator complex, as MED16 with WDR-deletion shows a dosage-dependent effect on its association with Mediator complex (as indicated by MED1 here); in details, this association gradually decreased as more WDR domains are truncated (Fig 2B). The C-terminal of MED16 harbors an αβ-domain with 157 amino acids (11, 48) and the MED16 truncated mutant without αβ-domain failed to interact with UBP1-TFCP2 (as indicated by TFCP2 here) (Fig 2B). Therefore, the N-terminal WDR domain of MED16 is crucial for its integration into the Mediator complex, while the C-terminal αβ-domain is essential for interacting with UBP1-TFCP2.

MED16 αβ-domain interacts with UBP1 specific 36 amino acids sequence in Vivo and in Vitro.

(A) Schematic of Flag-MED16 deletion mutants. The WDR domains are shown as blue boxes. “1” represent full-length MED16 and “2” to “9” represent different truncations of MED16. All proteins were labeled with Flag-tag and expressed in 293T cell line as panel (B) indicated. (B) Domain mapping experiment of MED16. Flag-MED16 deletion mutants were coimmunoprecipitated with endogenous MED1 and TFCP2 using anti-Flag antibody. Asterisks indicate the IgG heavy and light chains. (C) Purified GST and GST-MED16 αβ-domain proteins were incubated with 6×His-tag TFCP2, UBP1, TFCP2L1, and YY1 at 4℃. After washing the interaction protein were eluted by boiling and immunoblotted with antibodies against GST-tag and His-tag. (D) Amino acids sequence analysis of UBP1 and TFCP2 protein, a 36 amino acids UBP1-specific sequence (UBP1-SP) as shown below. (E) Schematic of Flag-UBP1 deletion mutants. FL represents full-length UBP1, △S represents SAM motif truncated UBP1, and DBD represents UBP1 DNA binding domain. (F) Domain mapping experiment of UBP1. Flag-UBP1 deletion mutants were coimmunoprecipitated with endogenous MED1, MED23, MED16, and TFCP2. Asterisks indicate the IgG heavy and light chains.

We conducted a GST pull-down experiment to assess the direct interaction between individual transcription factors and MED16. We expressed and purified UBP1, TFCP2, YY1, and TFCP2L1 that were fused with 6×His tag using bacterial (E.coli) expression system (Fig 2C). Transcription factor CP2 like 1 (TFCP2L1), a paralog of TFCP2, was included for comparison. Due the solubility issue with full length MED16, we purified only GST-fused C-terminal αβ-domain of MED16 (GST-M16αβ) that comprises 157 amino acids. GST-M16αβ exhibited the strongest binding to recombinant UBP1 (Fig 2C), suggesting that UBP1 binds to MED16 directly.

The amino acid sequences of UBP1 and TFCP2 share 88% identity, we investigated why UBP1, but not TFCP2, directly binds to MED16. Through comparison of the amino acid sequences of UBP1 and TFCP2, we found that a 36-amino-acid-long sequence between the DNA binding domain and SAM motif is specific to UBP1 (UBP1-SP) (Fig 2D). To test if UBP1-SP is responsible for interacting with MED16, a UBP1 domain-mapping experiment was performed (Fig 2E and 2F). The full-length Flag-tagged UBP1 successfully pulled down endogenous TFCP2 and MED16. In contrast, the SAM motif-deleted UBP1 truncated mutant could only bind to MED16 modestly, and failed to bind to TFCP2. DNA binding domain of UBP1 (DBD) didn’t bind either MED16 or TFCP2 at all. These results indicate that MED16 complexes with UBP1 and TFCP2 in vivo and in vitro, and that the UBP1-SP domain is crucial for the interaction with MED16.

MED16 is required for UBP1 activation of gene transcription

Because the interaction between transcription factors and Mediator subunits is critical for target gene activation, we analyzed if MED16 is required for UBP1-driven gene transcription using reporter gene assay (Fig 3A). In the UBP1 overexpression assay, the full-length UBP1 protein significantly promoted transcriptional activation. However, the DNA-binding domain of UBP1 (DBD, 1-274aa) alone was unable to activate the reporter. The truncated UBP1 mutant (ΔS, 1-369aa), which contains the MED16 binding motif but lacks the TFCP2-binding SAM motif, exhibited about half the transcriptional activation when compared to the full-length UBP1 (FL) (Fig 3B).

MED16 is required for UBP1 activation of gene transcription

UBP1 reporter assay (A) with co-transfection of UBP1, MED16, and their mutants in various cell lines. In HeLa cells (B), 300 ng of UBP1 DNA binding domain (DBD), SAM motif-deleted mutant (ΔS), or full-length UBP1 (FL) plasmid was transfected with the reporter. (C) In MED16 knockdown and control HeLa lines generated by shRNA, 300 ng of UBP1 plasmid or control vector was co-transfected with the UBP1 reporter. Knockdown efficiency was confirmed by qPCR and Western blotting (D). (E) In wild-type and MED16 knockout mouse pancreatic cancer HT cells, the UBP1 reporter was co-transfected with 300 ng of mouse UBP1 plasmid or vehicle control. Knockout efficiency was validated by Western blotting (F). (G) Effect of increasing MED16 doses (0.3 μg, 0.5 μg, 1 μg) or αβ-domain truncated MED16 on the UBP1 reporter. (H) Effects of MED16, UBP1, and TFCP2 combinations on UBP1 reporter activity. Firefly luciferase activity was normalized to Renilla luciferase activity. The luciferase expression levels between groups were compared using two-tail unpaired Student’s t-test, The data are presented as the means ± SDs from n = 3 independent experiments. *: p < 0.05, **: p-value<0.01 and “ns”: no significant.

To further confirm that MED16 is required for the transcriptional activity of UBP1, a HeLa cell line with MED16 knockdown was generated using viral-mediated shRNA targeting MED16, along with a negative control cell line (sh-NC). The UBP1-responsive reporter showed similar baseline of transcriptional activity in the two cell lines. The reporter was activated about 4-fold by overexpressing UBP1 in the negative control cell line, whereas knockdown of MED16 attenuated the activation to approximately two-fold (Fig 3C).

The coding sequences of human MED16 and Mouse Med16 share about 90% identity. Thus, we further established a Med16 knockout cell line generated using CRISPR-Cas9 gene editing method in a mouse pancreatic cancer cell line called HT. Five Med16 knockout clones (ko16a, ko16b, ko16c, ko16d, & ko16e), with different frameshift mutations were selected and verified by sequencing and Western-Blotting (Fig 3F). All Med16-KO clones displayed 2-7-fold reduced reporter activity under ectopic UBP1 activation when compared to the wildtype HT cells, and Med16 KO also exhibited decreased basal level of UBP1-reporter activity without UBP1 overexpression (Fig 3E and 3F). These results confirmed the specific requirement of MED16 for UBP1 transcriptional activation.

To further verify the specific regulation of UBP1 activity by MED16, we performed a titration experiment by co-transfected low and high doses of MED16 or MED16△αβ (Fig 2A, truncation “9”) with the UBP1-luciferase reporter. It has been reported that overexpression of Mediator subunits can result in inhibition of transcription owing to competition between the overexpressed subunits and the Mediator complex for the transcriptional activator (15, 49-52). In agreement with these results, UBP1 luciferase reporter activity tended to be increasingly repressed by MED16 in a dosage-dependent manner. In contrast, MED16△αβ, which cannot interact with UBP1, failed to repress the luciferase reporter activity (Fig. 3G). These findings highlight a specific coupling between MED16 and UBP1 in activating gene transcription.

The UBP1-TFCP2 hetero-dimerization is important for UBP1-responsive reporter transcription as it is shown above (Fig 3B). We then tested whether UBP1, TFCP2, and MED16 had a combinatorial effect on gene transcription. UBP1 and TFCP2 co-overexpression had better (from 9.8 times of basic level to 12 times of basic level) transcriptional activity than UBP1 alone, but MED16 co-overexpression reduced UBP1’s activity in all of our combinations (Fig 3H). Together, the results of the UBP1-responsive reporter suggested that MED16 is required for UBP1 activation of gene transcription.

Analysis of the UBP1 and MED16 co-regulated transcriptome

We went on to profile the UBP1 and MED16 co-regulated genes on a genome-wide scale using RNA-seq via comparison of the transcriptomes between MED16-knockdown and control HeLa cells with or without ectopic UBP1 expression. A total of 217 genes were activated (>1.5-fold), and 60 genes were repressed (>1.5-fold) in UBP1 overexpressing HeLa cells (Fig 4A). Among the 217 UBP1-activated genes, 198 were repressed by shMED16 knockdown (Fig 4B and 4C). This result demonstrated that UBP1 and MED16 collaborate to regulate transcription in a genome-wide scale, further supporting that MED16 is required for UBP1 transcription activities.

UBP1 and MED16 co-regulated transcriptome analysis

(A) The number of genes repressed or activated by UBP1 overexpression in the sh-NC and shMED16 HeLa cell line (FC > |1.5|, p ≤ 0.05). (B) Venn diagram of UBP1-activated genes in the sh-NC HeLa cell line and the shMED16 HeLa cell line (FC > |1.5|, p ≤ 0.05). 198 genes which were activated by UBP1 overexpression but abolished by sh-MED16 were defined as UBP1 and MED16 co-regulated genes. (C) Heatmap showing the RNA-seq analysis of 217 UBP1-activated genes in sh-MED16 and sh-NC cell line treated with or without UBP1 overexpression. UBP1-activated genes were defined in sh-NC cell line under UBP1 overexpression at fold change>|1.5|, p-value≤0.05. (D) GO analysis of 198 UBP1 and MED16 co-regulated genes. The top 10 biological process items were presented. (E) Volcano map analysis of differentially expressed genes in sh-NC HeLa cell line with or without UBP1 overexpression. Genes failed to be activated by UBP1 in the sh-MED16 HeLa cell line we’re indicated by "×". (F) UBP1 motif occurrence analysis of SFTPA1, SFTPA2, SFTPB and SFTPC. Motif occurrence were showed within the promoter region of the genes (-1000 bp to 200 bp relative to the TSS).

To explore the potential biological processes that are co-regulated by MED16 and UBP1, 198 co-regulated genes by MED16 and UBP1 were subjected to gene ontology (GO) analysis (Fig 4D and 4E). Interestingly, GO analysis revealed that the genes involved in the respiratory gaseous exchange process were regulated by MED16 and UBP1, such as SFTPA1, SFTPA2, SFTPB, and SFTPC (Fig 4E). These genes encode pulmonary surfactant associated proteins that promote alveolar stability by lowering the surface tension at the air-liquid interface in the peripheral air spaces. They are normally only expressed in alveolar epithelial cells and silenced in HeLa cells. However, highly matched UBP1 motifs were identified in their promoters (Fig 4F), leading to their activation when UBP1 was overexpressed. In our analysis, angiogenesis and cellular response to VEGF stimulus are also enriched in GO analysis (Fig 4D), which is consistent with the previous report that UBP1 plays a critical role in the regulation of extraembryonic angiogenesis (29, 53). Collectively, these results demonstrated that UBP1 and MED16 may collaborate to activate a target array of endogenous genes expression.

MED16 collaborates with UBP1 to inhibit HIV-1 transcription

Beyond their roles in transcriptional activation, UBP1 and TFCP2 have been reported to inhibit HIV-1 transcription through complex mechanisms (32-34). We questioned whether the MED16 also play a role in HIV-1 transcription repression. Thus, we constructed an HIV-1 reporter that consists of the HIV-1 core promoter (-78 to +60) and luciferase CDS (Fig 5A).

MED16 cooperates with UBP1 to inhibit HIV-1 reporter expression

(A) Schematic illustration of the HIV-1 reporter, with the HIV-1 core promoter and +1 to +60 TAR RNA sequence cloned upstream of the luciferase coding sequence (CDS). (B) The HIV-1 reporter was co-transfected with 1.5 μg of MED16, UBP1, TFCP2, YY1, or Flag-tag control plasmids in 293T cells. (C and D) Increasing amounts (0.7 μg, 1 μg, 1.5 μg) of MED16 or αβ-domain-deleted MED16 were transfected with the HIV-1 reporter. (E) The HIV-1 reporter was co-transfected with 1.5 μg of MED23, MED24, or Flag-tag control plasmids as additional controls. (F) Comparison of HIV-1 reporter luciferase activity between sh-NC and sh-MED16 293T cell lines. Firefly luciferase activity was normalized to Renilla luciferase activity. The luciferase expression levels between groups were compared using two-tail unpaired Student’s t-test, The data are presented as the means ± SDs from n = 3 independent experiment. * p < 0.05, ** p-value<0.01 and “ns”: no significant.

Previous research has shown that overexpression of UBP1 or YY1 inhibits HIV-1 transcription (34, 47). Consistently, we found our HIV-1 reporter’s activity was reduced by overexpressing either UBP1 or YY1 in HeLa cells, validating our reporter’s response to UBP1 and YY1 (Fig 5B). We then tested if MED16 participates the UBP1-driving HIV-1 transcription repression. We observed that overexpression of MED16 inhibited HIV-luciferase activity in a dose-dependent manner (Fig 5C). However, the MED16 mutant with the UBP1 interacting αβ-domain deleted failed to inhibit HIV-1 reporter despite similar increase in the dosage of transfected MED16△ αβ (Fig 5D). This result demonstrated that MED16-mediated HIV-1 inhibition depends on its binding to UBP1.

To see if this HIV-1 inhibition is MED16 specific, MED23 and MED24 that are part of the Mediator complex tail module neighboring MED16 were cotransfected with the HIV-1 reporter, respectively. Neither MED23 nor MED24 showed any effect on the activity of the HIV-1 reporter (Fig 5E), thus underscoring the specific role of MED16 in repressing HIV-1 transcription. Moreover, MED16 knockdown by sh-RNA also modestly increased HIV-1 luciferase activity (Fig 5F). Collectively, these results demonstrated that MED16 collaborates with UBP1 to inhibit HIV-1 transcription.

UBP1 binding site determines the UBP1-MED16 inhibition in HIV-1 transcription

In a previous study, UBP1 was shown to inhibit HIV-1 transcription by recruiting YY1 and HDAC1 (Coull et al., 2000). To investigate if MED16 employs a similar mechanism to repress HIV-1 transcription, we transfected HeLa cells with an HIV-1 reporter and treated them with panobinostat, a pan-HDAC inhibitor. Western blot analysis showed increased acetylation of histone H3 (H3ac) upon panobinostat treatment (Fig S2A). However, this treatment did not reverse MED16-mediated inhibition of HIV-driven luciferase activity (Fig S2A). These results suggest that MED16 does not rely on HDAC-mediated histone deacetylation to inhibit HIV-1 transcription and involves alternative mechanisms to repress HIV-1 transcription.

We then questioned whether the cis regulatory elements in the HIV-1 promoter are necessary for MED16-mediated inhibition. To address this question, we first examined whether the transactivating response region (TAR) plays a role in MED16-mediated HIV-1 inhibition HIV-1 inhibition. TAR is transcribed into TAR RNA, which recruits the host pTEFb-containing super elongation complex (SEC) (54-56) to mediate the activation of viral protein Tat (57). UBP1 has been reported to restrict HIV-1 transcription at the level of elongation (33). A TAR RNA mutation HIV-1 reporter was created by replacing TAR loop nucleotides 31-34 UGGG with CAAA, which reduces binding affinity of Tat:P-TEFb and abolishes the function of TAR RNA (58). We observed that MED16 inhibited both wild-type and TAR-mutant reporter’s transcription, while mutated TAR attenuated UBP1-mediated inhibition (Fig S2B and S2C). This result suggested that MED16-mediated inhibition of HIV-1 is independent of the TAR-mediated HIV-1 transcriptional elongation.

HIV-1 proximal promoter has a string of GC-box that is bound by SP1, and an AP4 binding site between TATA-box and TSS (59). We then examined whether these two transcription factors interact with MED16 for transcription repression. A reporter driven by 4×GC-box was co-transfected with the MED16 expression plasmid, but no significant inhibition was observed compared with co-transfecting with the control vector (Fig S2D), suggesting no cross-talk between MED16 and Sp1 (GC-box). Then we mutated the AP-4 binding motif CAGCTG to CAGTCG (Fig S2E) to abolish AP-4 binding (60). Overexpressing MED16 inhibited the luciferase activity, regardless AP-4 binding status, suggesting that MED16 and AP-4 don’t work together to inhibit the HIV-1 transcription.

We then tested if the UBP1-MED16 inhibition of HIV-1 transcription is due to the direct UBP1 binding at the transcription started site (TSS) (32, 33). To test this hypothesis, two chimera reporters were created. Both are driven by a chicken β-ACTIN promoter. We replaced the native TSS site of the first reporter (Fig 6A) with the HIV-1 TSS contains the UBP1 binding site, while the second (Fig 6B) featured a chicken β-ACTIN promoter and a standard UBP1 binding site placed upstream of it. Various truncated mutants of MED16 and UBP1 were co-transfected with these reporters, respectively. In the case of the first chimera reporter, transcription was inhibited by full-length MED16 and MED16-bound UBP1 variants, but not by UBP1-binding deficient MED16 or MED16-binding deficient UBP1. Conversely, transcription of the second chimera reporter was not inhibited by any MED16 or UBP1 variants (Fig 6B). Additionally, comparison of transcription activities between the two reporters revealed that the first chimera exhibited approximately 50% lower transcription activity than the second (Fig 6C). These findings suggest that the positioning of the UBP1 binding site could be critical for MED16-mediated transcriptional repression.

UBP1 binding site determines the UBP1-MED16 inhibition in HIV-1 transcription

Schematic showing a chimera reporter (A) driven by the Chicken β-ACTIN promoter, followed by an HIV-1 UBP1 binding site that also serves as the HIV-1 transcription start site and another chimera reporter (B) driven by the Chicken β-ACTIN promoter, a UBP1 binding site modify from HIV-1 TSS placed upstream of it. 1 μg of different truncation of MED16 or UBP1 plasmid was co-transfected with the reporter. The Luciferase activity comparison between two chimera reporters were shown in (C). “cβa-UBP1 bd” indicate the reporter that panel A described, and “UBP1 bd-cβa” indicate the reporter that panel B described. Firefly luciferase activity was normalized to Renilla luciferase activity. The luciferase expression levels between groups were compared using two-tail unpaired Student’s t-test, The data are presented as the means ± SDs from n = 3 independent experiment. * p < 0.05, ** p-value<0.01 and “ns”: no significant. UBP1 motif enrichment analysis of top 50 upregulated or top 50 downregulated transcripts induced by UBP1 were shown in (D). The transcripts changes levels are shown on the left as a heatmap, and average profile plots of UBP1 motif density are shown on the right. UBP1 motif occurrence analysis of UBP1 downregulated gene KIFC1 and GTF2H4 were shown in (E). Motif occurrence were showed within the promoter region of the genes (-500 bp to 200 bp relative to the TSS).

We then investigated whether the position-dependent function of UBP1 also influenced UBP1-regulated endogenous gene expression across the human genome. To explore this, we analyzed the correlation between the relative binding position of UBP1 to TSS and gene transcription levels. We selected the top 50 upregulated and top 50 downregulated transcripts under UBP1 overexpression and analyzed the enrichment of the UBP1 motif around their promoters. Notably, UBP1-activated transcripts showed significant enrichment of the UBP1 motif upstream and downstream of the TSS, while UBP1-repressed transcripts exhibited motif enrichment specifically overlapping the TSS (Fig 6D). Two UBP1-inhibited gene: KIFC1 and GTF2H4, were found to have their TSS overlapping with UBP1 binding motif (Fig 4E and 6E). Together, our data demonstrate that UBP1 may act as either a transcriptional activator or repressor, depending on distance of its binding site to TSS.

MED16-UBP1 complex prohibited PIC formation of HIV-1 transcription

We wondered how UBP1-MED16 binding within the proximal promoter might influence the HIV-1 transcription. To investigate this, we employed an in vitro immobilized template system. We first created nuclear extracts from three different cell lines derived from HT cells: specifically, wild type (WT), Med16 knockout (Med16 KO), and overexpression (Med16 OE). HIV-1 core promoter (-78 to +60) templates were labeled with a single biotin molecule at their 5’ ends and immobilized using streptavidin beads. These immobilized templates were then incubated with the NEs to form the pre-initiation complex (PIC). Transcription was initiated by adding four NTPs to stimulate transcription pause release and elongation. The binding proteins on the templates at each step were collected for western blotting, and the transcribed RNA was quantified (Fig 7A).

MED16-UBP1 complex prohibited PIC formation of HIV-1 transcription

(A) Schematic of biotin-labeled immobilized template assay. (B) Immobilized template assay in wild-type, Med16 knockout and Med16 overexpression HT cell line. The binding proteins in individual steps were collected for immunoblotting. Beads: the streptavidin beads incubated with NE as a negative control; PIC: the pre-initiation step of transcription; Trx.: transcription initiation from the PIC step by adding NTPs. (C) qPCR quantification of generated RNA from different NE. The mRNA expression levels were compared between groups using two-tail unpaired Student’s t-test, The values are presented as the means ± SDs from n = 3 independent experiment. ** p-value<0.01 and “ns”: no significant.

Med16 overexpression altered both PIC formation and transcription initiation stages of the immobilized template transcription process (Fig 7B). During the PIC forming stage, there was a decreased binding of CDK8, a key component of the Mediator kinase module. After transcription initiated, the signal for PolII pSer5, a well-known marker of transcription initiation, was reduced in MED16 OE NE compared to the WT (Fig 7B). Additionally, other Mediator subunits, such as MED12 from the kinase module, Med1 from the middle module, and MED23 from the tail module, exhibited decreased binding after transcription was initiated (Fig 7B). Moreover, Med16 overexpression prolonged the binding of general transcription factor TFIIB, which should be released after transcription initiation (61). Consequently, the Luciferase mRNA level in the MED16 OE was only 55% of that in the WT (Fig 7C). These results indicate that MED16 overexpression disrupts critical stages of HIV-1 transcription initiation, leading to a reduction in RNA production.

MED16 Maintains the HIV-1 latency

Because the MED16-UBP1 complex inhibits HIV-1 transcription (Figure 5), we explored whether it might also play a role in maintaining HIV-1 latency in CD4+ T cells. For this, we used the J-Lat 10.6 cell line, a Jurkat-based cell line infected with a pseudotyped HIV-1 strain (HIV/R7/E-/GFP) (Jordan et al., 2003). This cell line harbors the integrated HIV-1 copy in the second intron of the SEC16 gene (Chung et al., 2020) and the integrated HIV-1 copy remains silenced under normal conditions. HIV-1 latency reversal can be monitored by GFP expression that is controlled by viral promoter and serves as a marker for viral activation (Fig 8A). To induce viral expression, we treated cells with the BET bromodomain inhibitor JQ1 that is known to reverse HIV-1 latency by antagonizing BRD4’s suppression of the Tat-P-TEFb/SEC (Super Elongation Complex) interaction, (Li et al., 2013; Zhu et al., 2012).

MED16:UBP1 complex prohibited PIC formation of HIV-1 transcription

(A and B) Stable overexpression of MED16 (MED16 OE) in the J-Lat 10.6 cell line using MSCV retrovirus. Control: Control J-Lat 10.6 cell line generated via infection with a vehicle retrovirus. (C) Fluorescence microscopy images of MED16 OE and Control J-Lat 10.6 cells treated with or without 5 μM JQ1 for 48 hours. Percentage of GFP-positive cells measured by FACS (D), and mean fluorescence intensity (MFI) quantified using the BioTek Synergy 2 system (E). The values are presented as the means ± SDs from n = 3 independent experiment. ** p-value<0.01

We generated J-Lat 10.6 cell lines with retroviral MED16 expression and a control line infected with the MSCV (Murine Stem Cell Virus) retrovirus (Fig 8B). Prior to JQ1 treatment, both cell lines showed minimal GFP expression, confirming that MED16 manipulation alone did not alter HIV-1 latency (Fig 8C). After treatment with 5μM JQ1 for 48h, the HIV-LTR-driven GFP expression is observed in both cell lines, but is reduced in the MED16 OE cell line, with a 40% decrease in the percentage of GFP-positive cells percentage and 0.54-fold reduction in MFI (mean fluorescence intensity) compared to the control cell line (Fig 8C, 8D, and 8E). These findings suggest that MED16 not only inhibits HIV-1 transcription but also plays a crucial role in maintaining HIV-1 latency in CD4+ T cells.

Discussion

In this study we identified two transcription factors UBP1 and TFCP2 that interact with Mediator subunit MED16. This interaction activates an array of endogenous gene expression and inhibits HIV-1transcription.

Utilizing gel-filtration chromatography and immunoprecipitation-mass spectrometry (IP-MS), we found a significant amount of MED16 co-elutes at the fraction between 443 kDa and 150 kDa where other core Mediator subunits are absent. Several transcription factors, including TFCP2, UBP1, and YY1 were co-immunoprecipitated with MED16 within these fractions (Fig 1D, 1E and 1F). Although these proteins were isolated from the Mediator complex in gel-filtration chromatography, they functionally relied on the Mediator complex since the WDR-deleted MED16, which cannot interact with UBP1, failed to repress HIV-1 activation (Fig S3B). MED16 interacts with UBP1 to activate the expression of an array of endogenous genes. In the previous studies, UBP1 deficiency results in lethality at midgestation due to the prominent angiogenic defect observed in the mutant placentas and allantoic blood vessels fail to penetrate deeply and branch into the complex embryonic vasculature characteristic of the normal placenta (29, 53), which is consistent with our GO analysis that UBP1 and MED16 control angiogenesis gene expression. Unexpectedly, pulmonary surface molecule genes were also regulated by UBP1 and MED16. These gene encode pulmonary surfactant associated proteins, typically expressed in alveolar epithelial cells but silenced in HeLa cells. However, UBP1 overexpression activated them, likely due to the well-matched UBP1 motifs in their promoters.

Previous studies have demonstrated UBP1’s role in inhibiting HIV-1 transcription through various mechanisms, including epigenetic and elongational regulation. However, our investigation proposes a novel mechanism where the MED16-UBP1 complex inhibits HIV-1 transcription by preventing PIC formation at the HIV-1 TSS. This inhibition is specific to the position of the UBP1 binding site, with transcription being repressed when UBP1 binds to its site within a TSS and activated when UPB1 binds to its site either upstream or downstream of a TSS. This finding highlights the importance of binding site positioning in transcriptional control, which is often overlooked in past studies. The position-dependent function of transcription factors (TFs) relative to the TSS is supported by a recent study demonstrating how TF binding site’s spatial configuration can lead to distinct gene regulatory outcomes. (62) showed that the effect of TF binding on transcription initiation is highly position-dependent. TFs such as NRF1, NFY, and Sp1 activate or repress transcription depending on their exact position relative to the TSS. Our findings with the MED16-UBP1 complex align with this because UBP1 activates transcription when binding upstream or downstream of the TSS but represses it when binding directly at the TSS.

Targeted inhibition of HIV-1 transcription, particularly through enhancing the silencing activity of UBP1-MED16 at the viral transcription start site (TSS), presents a promising strategy for HIV-1 therapy. Many current approaches focus on inhibiting key regulatory proteins crucial for HIV-1 transcription (Ophinni et al., 2018; Peng et al., 2023; Yeh et al., 2020). By strengthening the UBP1-MED16 interaction or increasing MED16’s affinity for the HIV-1 TSS, it may be possible to achieve more effective and stable transcriptional silencing, which could reduce viral expression over prolonged periods. In the future, small molecule drug screening can be conducted targeting the enhancement of this interaction, and the obtained small molecules could be used in combination with existing antiretroviral therapies (ART) to enhance therapeutic efficacy (Pai et al., 2021; Sorokina et al., 2023).

Interestingly, a small peptide fragment of MED16 has the opposing effect of activating HIV-1 transcription while potentially blocking the endogenous UBP1-MED16 interaction (Fig S3C). By manipulating the expressing form of MED16, we can create a multi-faceted therapeutic tool that enables flexible regulation of HIV-1 transcription based on specific therapeutic needs. For instance, expressing the full-length MED16 could enhance the efficacy of antiretroviral therapies (ART), while expressing the MED16 αβ-domain peptide could support "shock-and-kill" strategies (63) by reactivating latent HIV-1, thereby enabling immune clearance or targeted antiretroviral intervention in previously hidden reservoirs. Together, these dual capabilities of MED16 position it as a versatile approach in the treatment of HIV-1.

Method

Plasmids

The human MED16, YY1, TFCP2, TFCP2L1 cDNAs were obtained from BRICS (Bio-Research Innovation center SUZHOU). The human UBP1 and mouse Med16 cDNAs were cloned from HeLa cells and HT cells respectively. Then full-length cDNA and truncations were cloned into 3×Flag-CMV-10 (Sigma-Aldrich-Aldrich) and pET-28b plasmids. The HIV-1 core promoter sequence were synthesized directly and clone into pGL3-Basic plasmid. All mutations on plasmid were generated by using FASR Site-Directed Mutagenesis Kit (Tiangen; Beijing; China). siMED16 and Ctrl shRNA oligonucleotides were cloned into pSiren-RetroQ (Clonetech).

Cell culture

HeLa and 293T cells were maintained in Dulbeccòs modified Eaglès medium (DMEM) containing 10% (v/v) FBS and 1% P/S. The HT cell line used in this study was an immortalized pancreatic cancer cell line derived from a KRASG12D P53+/- mouse and was generous provided by professor Xiaofei Yu in Fudan University. J-Lat 10.6 cell line was generous provided by professor Huanzhang Zhu in Fudan University. HT and J-Lat 10.6 cell were maintained in RPMI-1640 medium containing 10% (v/v) FBS and 1% P/S. All cells were maintained in the incubator with 5% CO2 at 37℃.

Western blot and real-time PCR assays

Method for Western blot and real-time PCR assay have been described previously (64). The primers for real-time PCR are listed in Table S1. Antibody for Western blot include the following MED23 (Abcam, ab200351), MED1 (Bethyl Lab, A300-793A), MED24 (Bethyl, A301-472A), MED6 (Santa Cruz, sc-9434), MED12 (Bethyl Lab, A300-774A), MED16 (Bethyl Lab, A303-668A), CDK8 (Abcam, ab115155), UBP1 (Protientech, 67318-1-Ig), TFCP2 (Proteintech, 15203-1-AP), YY1 (Proteintech, 66281-1-Ig), P300 (Abcam, ab14984), HDAC1 (Santa Cruz, sc-7872), PolII (Abcam, ab816), PolII-pSer5 (Millipore, 04-1572), TFIIB (Santa Cruz, sc-274D), TFIIH (Santa Cruz, sc-293), H3 (Abcam, ab1791), H3ac (Abcam, ab4729), β-Actin (Proteintech, 66009-1-Ig), GAPDH (Proteintech, 60004-1-Ig), Flag-tag (Sigma-Aldrich, F1804), His-tag (Proteintech, 66005-1-Ig), GST-tag (Invitrogen, 10004D).

Nuclear extract Preparation and gel-filtration chromatography assay

Preparation of un-dialyzed HeLa cell nuclear extract was described previously (65). Gel-filtration molecular weight markers (Sigma-Aldrich, MWGF1000) were loaded onto a Superose 6 10/300GL column to determine the void volume (Blue dextran ∼2000 kDa) and elution volume of protein standards (Thyoglobulin 669 kDa, Apoferritin 443kDa, Alcohol Dehydrogenase 150 kDa, Albumin 66 kDa). HeLa nuclear extract was applied to the Superose 6 10/300GL column and then was run in Buffer D (20 mM HEPES (pH 7.9), 300 mM KCl, 10% glycerol, 0.2 mM EDTA, 10 mM β-mercaptoethanol, and 1 mM PMSF). Fractions of 500µl were collected from the column and precipitated using trichloroacetic acid (TCA). Every third fraction was used for Western blot assay.

Co-immunoprecipitation (Co-IP) assay

For transient co-transfection, 293T cells plated in 10 cm dishes (90% confluency) were transfected with 10 μg of each plasmid using Lipofectamine 2000 (Invitrogen) according to the manufacturer’s protocol. After 48 h, the cells were harvested and lysed in lysis buffer (20mM HEPES pH7.5, 150mM NaCl, 1mM EDTA, 2.5mM EGTA, 0.3% Tween-20) with protease inhibitors. Lysates were subjected to immunoprecipitation with anti-Flag M2 beads (Sigma-Aldrich, A2220) and incubated at 4℃ overnight, following by washing in lysis buffer three times. Then 50 μl 1× SDS loading buffer was added to the beads and boiled at 99℃ for 10 min for Western blot analysis with the indicated antibodies.

For endogenous Co-IP, 293T cells were seeded in 10 cm dishes and cultured under appropriate conditions. At nearly 80% confluence, the cells were harvested and lysed in lysis buffer with protease inhibitors. Lysate was centrifuged at 4℃, 13200 rpm for 10 minutes, and the supernatant was incubated with 4μg antibodies to MED16 (Bethyl, A303-668A), CDK8 (CST, 17396s), MED1 (Bethyl, A300-793A), MED12 (Bethyl, A300-774A) overnight. Then, 20 μl of Dynabeads Protein G (Invitrogen) was added and incubated for 2 hours at 4℃. The beads were then washed with lysis buffer (with 0%, 5%, 10% 1,6-hexanediol) three times and boiled with SDS loading buffer for Western blot assay with indicated antibodies.

As for the Co-IP experiment in HeLa nuclear extracts (NE) or gel-filtration fractions, the HeLa NE was first diluted with an equal volume of D300 buffer containing 20 mM HEPES (pH 7.9), 300 mM KCl, 10% glycerol, 0.2 mM EDTA, 10 mM β-mercaptoethanol, and 1 mM PMSF. Next, 4 μg of antibody was added directly to the diluted HeLa NE or gel-filtration fractions, followed by incubation at 4°C overnight. After the overnight incubation, 20 μL of Dynabeads Protein G (Invitrogen) was added and incubated for an additional 2 hours at 4°C. The beads were then washed three times with D300 buffer and boiled in SDS loading buffer for Western blot analysis with the specified antibodies.

IP-MS and analysis

After immunoprecipitation, the protein samples were separated by SDS‒PAGE, running at 80 V for about 15 minutes. The electrophoresis was halted once the bromophenol blue dye front had traveled 0.5 cm into the resolving gel. The gel was then stained with Coomassie Brilliant Blue R250 to visualize protein bands, which were subsequently excised and diced into small fragments (1 mm³). These gel fragments underwent sequential washes: twice with a 50% acetonitrile (ACN) solution in 50 mM ammonium bicarbonate (NH4HCO3), followed by three washes with 10 mM NH4HCO3 and one wash with pure ACN to remove salts and detergents.

After dehydration with 100% ACN, the gel pieces were treated with 10 mM dithiothreitol (DTT) for 1 hour at 37 °C to reduce disulfide bonds. The DTT solution was then removed, and the gel fragments were dehydrated again with ACN. Next, 50 mM iodoacetamide (IAM) was added to alkylate free thiol groups, with incubation for 30 minutes in the dark. The gel pieces were washed three times with 10 mM NH4HCO3 and once with ACN before final dehydration.

For enzymatic digestion, the gel fragments were incubated overnight at 37 °C with trypsin (3 ng/μl in 10 mM NH4HCO3). The resulting peptides were extracted using a gradient ACN elution and analyzed by liquid chromatography-tandem mass spectrometry (LC-MS/MS) on an Orbitrap Fusion™ Lumos™ system.

Generation of Med16 KO cell line using CRISPR-Cas9

To generate Med16 knockout cell lines, the pX330-mCherry plasmid, containing the CRISPR-Cas9 machinery and target-specific guide RNAs, was used. A guide RNA sequence as reported in a previous study(66) was cloned into the pX330-mCherry vector. Cells were transfected with the pX330-mCherry plasmid using Lipofectamine 2000 (Invitrogen) according to the manufacturer’s protocol. After 48 hours, single mCherry-positive cells were sorted by fluorescence-activated cell sorting (FACS) into individual well of 96-wells plate. The sorted cells were then cultured and expanded. Individual clones were screened for gene knockout by sequencing and Western blotting to confirm successful editing.

Retrovirus infection

Stable cell lines with retroviral-mediated knockdown or overexpression were generated following the manufacturer’s protocol (Clontech) as described previously (13). Specific siRNA target sequences were designed using Thermo Scientific’s siRNA selection tools. Retroviruses were produced by co-transfecting 293T cells with pSiren-RetroQ (for knockdown) or pMSCV-puro (for overexpression) vectors along with the pCL10A1 helper plasmid, using Lipofectamine 2000 (Invitrogen). After 48 hours, the culture medium containing retroviruses was collected, supplemented with 20 μg/ml polybrene (Sigma-Aldrich-Aldrich), and filtered through a 0.45 μm filter. The filtered retroviruses were then added to target cells, which were subjected to spin infection at 2,500 rpm for 1.5 hours at 30℃. Twenty-four hours after infection, cells were selected with puromycin in 293T at 1 μg/ml and HeLa cells at 2.5 μg/ml.

Luciferase reporter assay

293T cells or HeLa cells were plated in a well of 12-wells plate (90% confluency) and transfected with 150 ng of luciferase reporter plasmid along with 50 ng pRL-TK (Promega) using Lipofectamine 2000 (Invitrogen). At 36-48h post transfection, the cells were harvested and subjected to dual-luciferase reporter assays according to the manufacturer’s protocol (Promega)

RNA-seq and data analysis

Total RNA was extracted from cells lysed with TRIZOL. We used a commercial RNA library kit to prepare the RNA-seq library according to the manufacturer’s specifications (VAHTS® Universal V8 RNA-seq LibraryPrep Kit) from Illumina. Briefly, mRNA was captured with mRNA capture beads before fragmentation. The RNA fragments were reverse transcribed and subjected to library amplification for Illumina (Novaseq 6000) sequencing.

For RNA-seq data analysis, the filtered clean reads were mapped to the human reference genome hg38 with HISAT2. Read counts for each gene were then quantified using HTSeq-count. Differential expression analysis was performed in R using DESeq2, which normalized the read counts and identified significantly differentially expressed genes.

Motif analysis

Motif enrichment analysis was performed using the closest known homolog of UBP1: the Tcfcp2l1 (GSE11431) position weight matrix (PWM) from the HOMER motif database (v4.9.1). Sequences and the Tcfcp2l1 PWM were submitted to the FIMO algorithm (MEME Suite v5.5.8) to scan for enriched motifs. This analysis generated base pair-resolution motif enrichment scores and associated p-values across all queried sequences.

In vitro transcription assays with immobilized template

The immobilized template assay was carried out based on the previous study(67). Biotin-labeled PCR products were generated from the HIV-1 reporter plasmid with biotin attached to the upstream end. These templates contain the HIV-1 core promoter region (-78 to +60) followed by the luciferase CDS. 6 μg of each biotinylated template was bound to 300 μg of Dynabeads M-280 Streptavidin (Dynal) according to the manufacturer’s instructions using a DynalMag™-2 (Dynal) for assistance. The immobilized templates were then resuspended in 100 μL of blocking buffer (20 mM HEPES, pH 7.9, 0.1 M KCl, 10% glycerol, 6 mM MgCl2, 2.5 mM DTT, 50 mg/mL BSA). After a 15-minute incubation at room temperature, 100 μL of nuclear extract from either WT, MED16-overexpressing, or Med16 KO HT cells was added and incubated at 30°C for 30 minutes to allow formation of pre-initiation complex (PIC). The immobilized templates were washed three times with buffer D containing 0.1% Triton X-100. The proteins bound to the templates were eluted with SDS gel loading buffer and analyzed by 10% SDS-PAGE.

For in vitro transcription, an NTP mix (25 mM each) was added to stimulate transcription after PIC formation, and the reaction was incubated at 30°C for 1 hour. The supernatant, containing the transcribed RNA products, was collected for qPCR quantification. The immobilized templates were again washed three times with buffer D plus 0.1% Triton X-100, and the proteins bound to the templates were eluted with SDS gel loading buffer and subjected to 10% SDS-PAGE.

Quantification and statistical analysis

For qPCR and luciferase assays, the means and SEMs were calculated from at least three independent experiments. Statistical significance was assessed using GraphPad Prism software, with data presented in graphs as means ± SDs. Student’s t-test was used for comparisons between two groups, with significance levels defined as *p < 0.05 and **p < 0.01.

eLife Assessment

The characterization of a dissociable Mediator subunit implicated in cellular pathways, particularly lung alveolar function and HIV latency, would be conceptually interesting. The authors have preliminary evidence for a stable Med16 subcomplex that may regulate specific genes. This work is useful in that it points to interactions between Med16 and UBP1, but the evidence is preliminary and incomplete.

Reviewer #1 (Public Review):

Summary:

Characterization of a dissociable Mediator subunit implicated in cellular pathways, particularly lung alveolar function, and HIV latency is conceptually interesting.

Strengths:

The strengths of this study are:

(1) Demonstration of MED16 dissociation from the core Mediator complex and formation of a subcomplex containing MED16, upstream-binding protein 1 (UBP1), and transcription factor cellular promoter 2 (TFCP2) by elegant biochemical fractionation and immunoblotting analysis.

(2) Defining nine N-terminal WD-40 repeats (WDRs) of MED16 as a Mediator-incorporating module and the C-terminal ⍺β-domain (157 amino acids) important for interaction with the UBP1-TFCP2 heterodimeric complex.

(3) Illustration of a weak hydrophobic interaction between MED16 and the Mediator core that could be disrupted by 1,6-hexanediol, but not by its 2,5-hexanediol isomer nor by high salt (500 mM NaCl) disruption.

(4) Classification of UBP1-upregulated cellular genes typically containing binding sites flanking the transcription start site (TSS) in contrast to UBP1-downregulated genes often containing a TSS-overlapping UBP1-binding site

(5) Presenting evidence for Mediator complex-dissociated free MED16-repressed HIV promoter activity through functional association with UBP1 and showing bromodomain-containing protein 4 (BRD4) inhibitor JQ1 that potentially disrupts BRD4-inhibited HIV-1 transcription elongation could lead to reversal of HIV-1 latency.

Weaknesses:

Nevertheless, foreseeable weaknesses include:

(1) No clear demonstration of MED16-UBP1-TFCP2 indeed forming a trimeric core subcomplex in regulating cellular gene transcription and HIV-1 promoter inhibition

(2) No validation of transcriptomic datasets and pathways identified.

(3) Use of mostly artificial reporter gene constructs and non-HIV host cells (e.g., human 293T embryonic kidney cells, human HeLa cervical cancer cells, and mouse HT pancreatic cancer cells) for examining MED16/UBP1-regulated HIV transcription.

(4) Inconsistent use of 293T and HeLa cells in the characterization of dissociated MED16 interaction with UBP1 and TFCP2.

(5) In vitro transcription using immobilized DNA templates was not performed to a high standard, thus failing to convincingly show MED16/UBP1-inhibited HIV-1 transcription preinitiation complex formation.

Reviewer #2 (Public Review):

Summary:

The article from Zheng et al. proposes an interesting hypothesis that the Med16 subunit of Mediator detaches from the complex, associates with transcription factor UBP1, and this complex activates or represses specific sets of genes in human cells. Despite my excitement upon reading the abstract, I was concerned by the lack of rigor in the experimental design. The only statement in the abstract that has some experimental support is the finding that Med16 dissociates from the Mediator and forms a subcomplex, but the data shown remain incomplete.

Strengths:

The authors have preliminary evidence that a stable Med16 complex may exist and that it may regulate specific sets of genes.

Weaknesses:

The experiments are poorly designed and can only infer possible roles for Med16 or UBP1 at this point. Furthermore, the data are often of poor quality and lack replication and quantitation. In other cases, key data such as MS results aren't even shown. Instead, we are given a curated list of only about 6 proteins (Figure S1), a subset of which the authors chose to pursue with follow-up experiments. This is not the expected level of scientific process.

(1) The data supporting the Med16 dissociation and co-association with UBP1 are incomplete and not convincing at this stage. According to the Methods and text, the gel filtration column was run with "un-dialyzed HeLa cell nuclear extract" and eluted in 300mM KCl buffer. The extracts were generated with the Dignam/Roeder method according to the text. Undialyzed, that means the extract would be between 0.4 - 0.5M NaCl. Under these high salt conditions (not physiological), it's possible and even plausible that Mediator subunits could separate over time. This caveat is not mentioned or controlled for by the authors. Because a putative Med16 subcomplex is a foundational point of the article, this is concerning.

The data are incomplete because a potential Med16 complex is not defined biochemically. The current state suggests a smaller Med16-containing complex that may also contain UBP1 and other factors, but its composition is not determined. This is important because if you're going to conclude a new and biologically relevant Med16 complex, which is a point of the article, then readers will expect you to do that.

Equally concerning are the IP-western results shown in Figure 1. In my opinion, these experiments do nothing to support the claims of the authors. The authors use hexanediols at 5% or 10% in an effort to disrupt the Mediator complex. Assuming this was weight/volume, that means ~400 to 800mM hexanediol solution, which is fairly high and can be expected to disrupt protein complexes, but the effects haven't been carefully assessed as far as I'm aware. The 2,5 HD (Figure 1B) experiments appear to simply contain greater protein loading, and this may contribute to the apparent differential results. In fact, in looking at the data, it seems that all MED subunits probed show the same trend as Med16. They are all reduced in the 1,6HD experiment relative to the 2,5 HD experiment. But it's hard to know, because replicates weren't completed and quantitation was not done. There aren't even loading controls. Other concerns about the IP-Western experiments are outlined in point 2.

(2) At no point do the authors apply rigorous methods to test their hypothesis. Instead, methods are applied that have been largely discredited over time and can only serve as preliminary data for pilot studies, and cannot be used to draw definitive conclusions about protein function.

a) IP-westerns are fraught with caveats, especially the way they were performed here, in which the beads were washed at relatively low salt and then eluted by boiling the beads in loading buffer. This will "elute" bound proteins, but also proteins that non-specifically interact with or precipitate on the beads. And because Westerns are so sensitive, it is easy to generate positive results. It's just not a rigorous experiment.

b) Many conclusions relied on transient transfection experiments, which are problematic because they require long timeframes, during which secondary/indirect effects from expression/overexpression will result. This is especially true if the proteins being artificially expressed/overexpressed are major transcription regulators, which is the case here. It is simply impossible to separate direct from indirect effects with these types of experiments. Another concern is that there was no effort to assess whether the induced protein levels were near physiological levels. Protein overexpression, especially if the protein is a known regulator of pol2 transcription (e.g., UBP1 or Med16), will create many unintended consequences.

c) Many conclusions were made based upon shRNA knockdown experiments, which are problematic because they require long timeframes (see above point), which makes it nearly impossible to identify effects that are direct vs. indirect/secondary/tertiary effects. Also, shRNA experiments will have off-target effects, which have been widely reported for well over a decade. An advantage of shRNA knockdowns is that they prevent genetic adaptation (a caveat with KO cell lines). A minimal test would be to show phenotypic rescue of the knockdown by expressing a knockdown-resistant Med16 (for example), but these types of experiments were not done.

d) Many experiments used reporter assays, which involved artificial, non-native promoters. Reporters are good for pilot studies, but they aren't a rigorous test of direct regulatory roles for Med16 or other proteins. Reporters don't even measure transcription directly. In fact, no experiment in this study directly measures transcription. An RNA-seq experiment was done with overexpressed or Med16 knockdown cells, but these required long timeframes and RNA-seq measures steady-state mRNA, which doesn't test the potential direct effects of these proteins on nascent transcription.

e) The MS experiments show promise, but the data were not shown, so it's hard to judge. The reader cannot compare/contrast the experiments, and we have no indication of the statistical confidence of the proteins identified. How many biological replicate MS experiments were performed?

(3) The data are over-interpreted, and alternative (and more plausible) hypotheses are ignored. Many examples of this, some of which are alluded to in the points above. For example, Med16 loss or overexpression will cause compensatory responses in cells. An expected result is that Mediator composition will be disrupted, since Med16 directly interacts with several other subunits. Also in yeast, the Robert, Gross, and Morse labs showed that loss of Med16/Sin4 causes loss of other tail module subunits, and this would be expected to cause major changes in the transcriptome. The authors also mention that yeast Med16/Sin4 "alters chromatin accessibility globally" and this would be expected to cause major changes in the transcriptome, leading to unintended consequences that will make data analysis and identification of direct Med16 effects impossible. The unintended consequences will be magnified with prolonged disruption of MED16 levels in cells (e.g., longer than 4h). These unintended consequences are hard to predict or define, and are likely to be widespread given the pivotal role of Mediator in gene expression. One unintended consequence appears to be loss of pol2 upon Med16 over-expression, as suggested by the western blot in Figure 8B. I point this out as just one example of the caveats/pitfalls associated with long-term knockdowns or over-expression.

Reviewer #3 (Public Review):

Summary:

There are two major flaws that fundamentally undermine the value of the study. First, nearly all the central conclusions drawn here rely on the unfounded assumption that the effects observed are direct. No rigorous cause-and-effect relationships are established to support the claims. Second, the quality of the experimental data is substandard. Collectively, these concerns significantly limit any advances that might be gained in our understanding of the UBP1 pathway or Mediator function.

Weaknesses:

(1) The decrease in 1,6-hexanediol-treated cells of MED16 is modest, variable, not quantified, and internally inconsistent. For example, in Figure 1A, 1,6-hexanediol treatment should not have an impact on the level of the protein being directly IP. For MED12 (and CDK8 and MED1 to a lesser extent), 1,6-hexanediol treatment alters the level of the target protein in the IP. Along these lines, Figure 1A shows a no 1,6H-D dependent decrease in MED1 or MED12 levels in the CDK8 IP, whereas Figure 1B does show a decrease. Figure 1A shows no 1,6H-D dependent decrease in CDK8 levels in the MED1 IP, whereas Figure 1B shows a dramatic decrease. MED24 levels in the MED12 IP increase upon 1,6H-D in Figure 1A, but decrease in Figure 1B. Internal inconsistencies of this nature persist in the other Figures.

(2) Undermining the value of Figure 1E/F, UBP1 and TFCP2 may also associate with the small amount of MED16 in the 2MDa fractions. This is not tested, and therefore, the conclusion that they just associate with the dissociable form of MED16 is not supported.

(3) Domain mapping studies in Figure 2 are overinterpreted. Since the interactions could be indirect, it is not accurate to conclude "Therefore, the N-terminal WDR domain of MED16 is crucial for its integration into the Mediator complex, while the C-terminal αβ-domain is essential for interacting with UBP1-TFCP2. "

(4) A close examination of Figure 2C undermines confidence in the association studies. The bait protein in lanes 5-8 should be equal. Also, there is significant binding of GST to UBP1 and TFCP2, in roughly the same patterns as they bind to GST-MED16 αβ. The absence of input samples makes the results even more difficult to interpret.

(5) The domain deletion mutants are utilized throughout the manuscript as evidence of the importance of the UBP1-MED16 interaction. However, in Figure 2F lanes 7 and 8, the delta-S mutant binds MED16 as well as full-length UBP1. This undermines much of the subsequent data and conclusions about specificity.

(6) Even if the delta-S mutant were defective for MED16 binding, the result in Figure 3B does not "confirm that MED16 is required for the transcriptional activity of UBP1,". Removal of that domain may have other effects.

(7) As Mediator is critical for the activation of many genes, it is not accurate to assume that the impact of its deletion in Figure 3E/F demonstrates a direct requirement in UBP1-driven transcription. This could easily be an indirect effect.

(8) Without documenting the relative protein expression levels in Figure 3G/H, conclusions cannot be drawn about the titration experiments, nor the co-expression experiments. These findings are likely the result of squelching or some form of competition that is not directly related to the UBP1-mediated transcription. A great deal of validation would be required in order to support the model that these effects are a result of MED16 overexpression sequestering UBP1 away from holo-Mediator.

(9) The lack of any documentation of expression levels for the various ectopic proteins in the majority of Figures, renders mechanistic claims meaningless (Figures 3, 4, 5, 6, 7, S2, S3). This is particularly relevant since the model presented for many of the results invokes concentration-dependent competition.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

Characterization of a dissociable Mediator subunit implicated in cellular pathways, particularly lung alveolar function, and HIV latency is conceptually interesting.

Strengths:

The strengths of this study are:

(1) Demonstration of MED16 dissociation from the core Mediator complex and formation of a subcomplex containing MED16, upstream-binding protein 1 (UBP1), and transcription factor cellular promoter 2 (TFCP2) by elegant biochemical fractionation and immunoblotting analysis.

(2) Defining nine N-terminal WD-40 repeats (WDRs) of MED16 as a Mediator-incorporating module and the C-terminal ⍺β-domain (157 amino acids) important for interaction with the UBP1-TFCP2 heterodimeric complex.

(3) Illustration of a weak hydrophobic interaction between MED16 and the Mediator core that could be disrupted by 1,6-hexanediol, but not by its 2,5-hexanediol isomer nor by high salt (500 mM NaCl) disruption.

(4) Classification of UBP1-upregulated cellular genes typically containing binding sites flanking the transcription start site (TSS) in contrast to UBP1-downregulated genes often containing a TSS-overlapping UBP1-binding site

(5) Presenting evidence for Mediator complex-dissociated free MED16-repressed HIV promoter activity through functional association with UBP1 and showing bromodomain-containing protein 4 (BRD4) inhibitor JQ1 that potentially disrupts BRD4-inhibited HIV-1 transcription elongation could lead to reversal of HIV-1 latency.

Weaknesses:

Nevertheless, foreseeable weaknesses include:

(1) No clear demonstration of MED16-UBP1-TFCP2 indeed forming a trimeric core subcomplex in regulating cellular gene transcription and HIV-1 promoter inhibition

We appreciate this insightful comment from the reviewer. The physical interaction between MED16 and the UBP1–TFCP2 heterodimer was uncovered serendipitously via biochemical fractionation. Specifically, UBP1 and TFCP2 were identified in gel-filtration fractions containing free MED16 that eluted separately from the intact core Mediator complex (Figures 1D–1F). This biochemical observation suggested to us a potential functional linkage between MED16 and these two transcription factors. Subsequently, our functional assays further clarified this regulatory relationship: UBP1 and TFCP2 do interact with MED16 that is required to be fully assembled into Mediator complexes to exert transcriptional control. Truncation of MED16’s N-terminal WD40-repeat (WDR) domain (MED16ΔWDR) eliminates MED16 integration into the Mediator complex. Notably, this MED16ΔWDR mutant still retains an intact C-terminal αβ-domain capable of binding UBP1 yet completely loses the capacity to support UBP1-dependent transcription. Collectively, these data demonstrate that UBP1–TFCP2 engage MED16 as a component of the complete Mediator complex to govern target gene expression in cells.

We have updated the abstract and full manuscript to present this model more clearly and coherently throughout the text.

(2) No validation of transcriptomic datasets and pathways identified.

We thank the reviewer for this comment. Since our RNA-seq analysis identified SFTP family genes (SFTPA1, SFTPA2, SFTPB, SFTPD) as UBP1-MED16 co-regulated targets, and these are lung-specific surfactant protein genes, we have now performed qRT-PCR validation in the A549 human lung adenocarcinoma cell line. The results confirm that knockdown of MED16 largely attenuates the expression of these SFTP genes (see Author response image 1) and other UBP1 target genes identified in our RNA-seq dataset.

qRT-PCR validation of SFTP gene family expression (SFTPA1/A2/B/D) in A549 cells

(3) Use of mostly artificial reporter gene constructs and non-HIV host cells (e.g., human 293T embryonic kidney cells, human HeLa cervical cancer cells, and mouse HT pancreatic cancer cells) for examining MED16/UBP1-regulated HIV transcription.

To overcome the limitations of artificial reporter systems, we have managed to utilize multiple complementary approaches beyond reporter assays, including: (i) RNA-seq analysis of endogenous gene regulation, (ii) biochemical co-IP and gel filtration experiments, (iii) immobilized template transcription assays that directly measure PIC assembly and nascent RNA production, and (iv) validation in the J-Lat 10.6 CD4+ T cell line harboring an integrated latent HIV-1 provirus, where MED16 overexpression suppresses HIV-1 reactivation.

Unfortunately, the direct infection experiments with replication-competent HIV-1 were not feasible as our institution does not have the biosafety level 3 (BSL-3) containment facilities.

(4) Inconsistent use of 293T and HeLa cells in the characterization of dissociated MED16 interaction with UBP1 and TFCP2.

For biochemical characterization in this study, HEK293T cells were primarily utilized for overexpression-based co-immunoprecipitation (Co-IP) and domain-mapping assays owing to their superior transfection efficiency. In contrast, HeLa cells were employed for gel-filtration chromatography and endogenous Co-IP, as fractionation of HeLa nuclear extracts represents a well-established workflow for purifying native protein complexes. Critically, the physical interaction between MED16 and the UBP1–TFCP2 heterodimer was consistently detected in both HEK293T whole-cell lysates and HeLa nuclear extracts across all assays.

(5) In vitro transcription using immobilized DNA templates was not performed to a high standard, thus failing to convincingly show MED16/UBP1-inhibited HIV-1 transcription preinitiation complex formation.

We appreciate the reviewer’s careful assessment of the immobilized template assay. While we recognize the technical challenges inherent in this approach, we want to emphasize that the proposed model—that MED16-UBP1 inhibits HIV-1 transcription at the level of PIC assembly—does not rest solely on this single experiment.

Multiple independent lines of evidence converge on this model:

(1) The MED16-UBP1 inhibition depends on the UBP1 binding site overlapping the TSS: a chimeric reporter with the UBP1 binding site placed at the TSS is inhibited by MED16, whereas placing it upstream does not show inhibition, demonstrating the UBP1 binding site-specific repression (Fig 6A-C).

(2) MED16 overexpression inhibits HIV-1 reporter activity in a manner depending on interaction with UBP1. Deletion of the MED16 αβ-domain (MED16Δαβ), which abolishes MED16-UBP1 interaction, completely abrogates the inhibitory effect (Fig 5E-G).

(3) The MED16-UBP1 inhibition of HIV is independent of TAR-mediated elongation (Fig S2B-C) and HDAC-mediated deacetylation (Fig S2A), consistent with an inhibition mechanism operating at the initiation stage rather than at the post-initiation steps.

(4) Upon MED16 overexpression, the immobilized template assay shows reduced CDK8 binding and Pol II pSer5 signals, along with TFIIB retention, providing mechanistic evidence that is consistent with the aforementioned functional observations.

Thus, the model is supported by convergent evidence from reporter assays, domain-mapping experiments, pharmacological inhibition, and in vitro biochemical reconstitution. In the revised manuscript, we have further clarified how these complementary approaches collectively support the PIC inhibition model.

Reviewer #2 (Public review):

Summary:

The article from Zheng et al. proposes an interesting hypothesis that the Med16 subunit of Mediator detaches from the complex, associates with transcription factor UBP1, and this complex activates or represses specific sets of genes in human cells. Despite my excitement upon reading the abstract, I was concerned by the lack of rigor in the experimental design. The only statement in the abstract that has some experimental support is the finding that Med16 dissociates from the Mediator and forms a subcomplex, but the data shown remain incomplete.

We would like to first address a core conceptual point that we believe underpins several of the reviewer’s concerns.

We respectfully disagreed that the reviewer summarized our proposed model as a pathway where “Med16 detaches from the complex, associates with transcription factor UBP1, and this complex activates or represses specific sets of genes.” We acknowledge that phrasing in our original manuscript may have created this misleading interpretation, yet this is not the mechanistic model we intend to convey. We identified the MED16–UBP1–TFCP2 interaction serendipitously during biochemical fractionation: UBP1 and TFCP2 co-eluted with free MED16 in gel-filtration fractions that were fully separated from the intact core Mediator complex. While this biochemical result guided our subsequent functional investigation into MED16–UBP1–TFCP2 crosstalk, we never intend to propose that a standalone MED16–UBP1–TFCP2 subcomplex can drive transcriptional activation or repression independently of the intact Mediator complex.

Most importantly, our functional experiments demonstrate that UBP1 and TFCP2 require MED16 as an integral part of the Mediator complex to regulate transcription as shown by domain mapping experiments. Specifically, MED16 uses its N-terminal WDR domain to integrate into the Mediator and its C-terminal αβ-domain to bind UBP1/TFCP2. Deletion of either WDR or αβ-domain is functionally inert in both cellular and HIV-1 reporter assays. Thus, the MED16-UBP1 interaction is functionally relevant because it links these transcription factors to the Mediator complex, not because it forms an autonomous regulatory complex. In another word, UBP1–TFCP2 target MED16 to recruit the whole Mediator complex for governing gene expression in cells.

We have comprehensively revised the full manuscript to sharply distinguish our initial biochemical observation from the validated functional model. As all experimental approaches carry inherent limitations, we tried our best to validate every major conclusion using multiple orthogonal assays and to draw mechanistic interpretations when results from distinct experimental systems do converge. Our detailed replies to each specific concern from the reviewer are provided below.

Strengths:

The authors have preliminary evidence that a stable Med16 complex may exist and that it may regulate specific sets of genes.

Weaknesses:

The experiments are poorly designed and can only infer possible roles for Med16 or UBP1 at this point. Furthermore, the data are often of poor quality and lack replication and quantitation. In other cases, key data such as MS results aren't even shown. Instead, we are given a curated list of only about 6 proteins (Figure S1), a subset of which the authors chose to pursue with follow-up experiments. This is not the expected level of scientific process.

As clarified in our Summary Response above, we thought this reviewer may have misinterpreted our data and model. Overall, our work demonstrates that MED16 acts as a molecular bridge, linking UBP1/TFCP2 to the Mediator complex, not that three proteins form a free-standing subcomplex, and we have extensively revised the manuscript to clarify this.

Here we also want to address the specific concerns about data completeness and rigor. The full mass spectrometry datasets were not clearly presented in the original submission. We have now reanalyzed the IP-MS data using a more precise quantification approach. Rather than scoring proteins by simple presence or absence, we calculated a bait-preferential score for each protein by comparing its normalized abundance in the MED16 IP to that in the other three Mediator bait IPs (MED1, MED12, CDK8). This score reflects the degree to which a protein is preferentially enriched by MED16 relative to other Mediator subunits. The bait-specific candidates identified by this analysis are presented in Fig S1B, and the complete list of MED16-specific candidates is now included. In addition, the raw abundance values and peptide counts for all identified proteins have been compiled and uploaded as supplementary tables. A detailed description of the normalization and scoring procedure is provided in the Methods section

Furthermore, we have added quantification analysis for key Western blot results throughout the revised figures.

(1) The data supporting the Med16 dissociation and co-association with UBP1 are incomplete and not convincing at this stage. According to the Methods and text, the gel filtration column was run with "un-dialyzed HeLa cell nuclear extract" and eluted in 300mM KCl buffer. The extracts were generated with the Dignam/Roeder method according to the text. Undialyzed, that means the extract would be between 0.4 - 0.5M NaCl. Under these high salt conditions (not physiological), it's possible and even plausible that Mediator subunits could separate over time. This caveat is not mentioned or controlled for by the authors. Because a putative Med16 subcomplex is a foundational point of the article, this is concerning.

The data are incomplete because a potential Med16 complex is not defined biochemically. The current state suggests a smaller Med16-containing complex that may also contain UBP1 and other factors, but its composition is not determined. This is important because if you're going to conclude a new and biologically relevant Med16 complex, which is a point of the article, then readers will expect you to do that.

We appreciate the reviewer’s thorough feedback and address the two core concerns separately below.

(1) Undialyzed nuclear extracts & high-salt conditions

Published work confirms Mediator undergoes massive precipitation (~70% loss) upon dialysis to low-salt 100 mM KCl buffer (1), justifying our use of undialyzed HeLa nuclear extracts for biochemical assays. Prior study also demonstrated intact Mediator remains stable in up to 1 M KCl (1). Therefore, MED16 dissociation seen at 300 mM KCl arises from its unique biochemical property, rather than non-specific high-salt damage.

(2) Biochemical interpretation of MED16-containing fractions

We have revised ambiguous original wording to clarify our model: we do not intend to propose an autonomous MED16–UBP1–TFCP2 subcomplex functions independently in cells. Gel filtration only captures a biochemical state where MED16 co-elutes with UBP1/TFCP2 separate from core Mediator, which merely inspired our functional follow-up. Three lines of evidence prove UBP1/TFCP2 require fully assembled Mediator to function:

(i) MED16’s N-terminal WDR domain mediates Mediator integration, while its C-terminal αβ-domain binds UBP1; both interactions can occur simultaneously (Fig 2B).

(ii) MED16Δαβ incorporates into Mediator but loses UBP1 binding, showing no activity in UBP1-driven and HIV-1 reporter assays (Fig 3G, 5G).

(iii) MED16ΔWDR retains UBP1 binding yet cannot assemble into Mediator, thus fails to repress HIV-1 transcription (Fig 5F).

In vitro MED16 separation only reflects its extractable biochemical behavior, not a functional standalone complex. Collectively, our data establish UBP1/TFCP2 recruit the entire Mediator complex via MED16. All relevant manuscript sections have been rewritten to distinguish this preliminary biochemical observation from our validated functional model.

Equally concerning are the IP-western results shown in Figure 1. In my opinion, these experiments do nothing to support the claims of the authors. The authors use hexanediols at 5% or 10% in an effort to disrupt the Mediator complex. Assuming this was weight/volume, that means ~400 to 800mM hexanediol solution, which is fairly high and can be expected to disrupt protein complexes, but the effects haven't been carefully assessed as far as I'm aware. The 2,5 HD (Figure 1B) experiments appear to simply contain greater protein loading, and this may contribute to the apparent differential results. In fact, in looking at the data, it seems that all MED subunits probed show the same trend as Med16. They are all reduced in the 1,6HD experiment relative to the 2,5 HD experiment. But it's hard to know, because replicates weren't completed and quantitation was not done. There aren't even loading controls. Other concerns about the IP-Western experiments are outlined in point 2.

Treatment with 10% (w/v) 1,6-hexanediol (1,6-HD) is a well-established method to selectively disrupt weak hydrophobic interactions while maintaining stable protein assemblies. A comparable strategy was recently applied to dissect NEAT1-dependent paraspeckle formation driven by liquid–liquid phase separation (2). Notably, we included 2,5-hexanediol (2,5-HD) as a critical specificity control: as a structural isomer with distinct hydroxyl group positioning, 2,5-HD cannot efficiently disrupt weak hydrophobic contacts and failed to trigger MED16 dissociation (Figure 1B). The divergent outcomes observed side-by-side with equal sample input therefore serve as an internal control, validating that MED16 dissociation is a specific consequence of 1,6-HD treatment.

We acknowledge mild signal reductions for several other Mediator subunits following 1,6-HD exposure under certain immunoprecipitation (IP) conditions (Figures 1A, 1B). Nevertheless, MED16 exhibits a uniquely pronounced dissociation by 1,6-HD that is consistent across three independent IPs using anti-CDK8, anti-MED1, and anti-MED12 antibodies, in both HEK293T whole-cell lysates and HeLa nuclear extracts. By contrast, the mild depletion of other Mediator subunits varies substantially depending on the antibody and cell extract used, whereas robust MED16 loss is universally observed in all experimental setups we tested.

(2) At no point do the authors apply rigorous methods to test their hypothesis. Instead, methods are applied that have been largely discredited over time and can only serve as preliminary data for pilot studies, and cannot be used to draw definitive conclusions about protein function.

We greatly appreciate the reviewer’s emphasis on methodological rigor. We fully concur that individual techniques including protein fractionation, IP-WB, transient transfection, shRNA knockdown, immobilized template and luciferase reporter assays each carry intrinsic limitations when deployed alone. To address this, we employed a panel of orthogonal experimental strategies throughout our work and only draw conclusions supported by convergent results across distinct assays. Our core mechanistic conclusion—that MED16 acts as a molecular bridge linking the UBP1–TFCP2 heterodimer to the intact Mediator complex to control transcription—is validated by multiple independent lines of evidence, rather than resting on a single experimental approach, as elaborated in detail below.

(a) IP-westerns are fraught with caveats, especially the way they were performed here, in which the beads were washed at relatively low salt and then eluted by boiling the beads in loading buffer. This will "elute" bound proteins, but also proteins that non-specifically interact with or precipitate on the beads. And because Westerns are so sensitive, it is easy to generate positive results. It's just not a rigorous experiment.

We acknowledge that IP-Western has inherent limitations regarding specificity. To minimize non-specific signals, we included IgG controls in our CDK8 IP experiments (available in raw data) and the boiling elution method has been extensively validated in our hands. More importantly, the interaction between MED16 and UBP1/TFCP2 is not solely supported by IP-Western.

This interaction was initially observed through co-elution of MED16 with UBP1 and TFCP2 in gel filtration fractions (Fig 1D), confirmed by reciprocal co-IP with anti-MED16 and anti-TFCP2 antibodies in both HeLa NE and the resolved fractions (Fig 1E-F), and further corroborated by in vitro GST pull-down assays showing direct binding between MED16 αβ-domain and recombinant UBP1 (Fig 2C). IP-MS analysis with three biological replicates independently identified UBP1 and TFCP2 as MED16-associated proteins. Each of these methods has different sources of potential artifacts, yet they all converge on the same conclusion.

(b) Many conclusions relied on transient transfection experiments, which are problematic because they require long timeframes, during which secondary/indirect effects from expression/overexpression will result. This is especially true if the proteins being artificially expressed/overexpressed are major transcription regulators, which is the case here. It is simply impossible to separate direct from indirect effects with these types of experiments. Another concern is that there was no effort to assess whether the induced protein levels were near physiological levels. Protein overexpression, especially if the protein is a known regulator of pol2 transcription (e.g., UBP1 or Med16), will create many unintended consequences.

The reviewer raises a valid concern about indirect effects from overexpression. To address this, our key conclusions were validated by multiple loss-of-function approaches that do not rely on overexpression alone. Specifically, the conclusion that MED16 is required for UBP1 transcriptional activity is supported by:

(i) shRNA-mediated MED16 knockdown (Fig 3C);

(ii) CRISPR-mediated MED16 knockout in five independent frameshift clones with consistent phenotypes (Fig 3E-F);

(iii) Using a domain-disruption approach, we demonstrated that a UBP1-binding-deficient MED16 mutant (Δαβ) fails to exert the dominant-negative effect on UBP1 activity, as seen with wild-type MED16 (Fig 3G and 6A). All these different methods point to the same conclusion. This makes it unlikely that the effects we see are simply artifacts of overexpression or indirect consequences.

(c) Many conclusions were made based upon shRNA knockdown experiments, which are problematic because they require long timeframes (see above point), which makes it nearly impossible to identify effects that are direct vs. indirect/secondary/tertiary effects. Also, shRNA experiments will have off-target effects, which have been widely reported for well over a decade. An advantage of shRNA knockdowns is that they prevent genetic adaptation (a caveat with KO cell lines). A minimal test would be to show phenotypic rescue of the knockdown by expressing a knockdown-resistant Med16 (for example), but these types of experiments were not done.

We agree that shRNA has potential off-target effects, and that is why we did not rely on shRNA as our sole loss-of-function approach. We have generated five independent CRISPR MED16 knockout clones, each with a distinct frameshift mutation, and all five clones showed consistent reduction in UBP1 reporter activity (Fig 3E-F). Furthermore, we performed RNA-seq on MED16 knockout cells in which wild-type MED16 was re-introduced. As presented in Author response image 2, re-expressing MED16 in KO cells can largely restore the expression of UBP1 target genes. These rescue data provide strong evidence that the transcriptional defects are specifically attributable to MED16 loss rather than off-target effects or clonal adaptation.

Gene profiling of MED16 KO and rescue effect

(d) Many experiments used reporter assays, which involved artificial, non-native promoters. Reporters are good for pilot studies, but they aren't a rigorous test of direct regulatory roles for Med16 or other proteins. Reporters don't even measure transcription directly. In fact, no experiment in this study directly measures transcription. An RNA-seq experiment was done with overexpressed or Med16 knockdown cells, but these required long timeframes and RNA-seq measures steady-state mRNA, which doesn't test the potential direct effects of these proteins on nascent transcription.

The two principal reporters used in this study are both based on native promoters. The UBP1 reporter is derived from the mouse α-globin gene promoter, a well-established natural target of UBP1/TFCP2 that contains two endogenous UBP1 binding sites. To enhance sensitivity, we expanded these to six copies, as the α-globin promoter has low basal activity in most cell lines. Critically, this expansion preserved the original promoter architecture, including the CCAAT box, TATA box, and TSS. The HIV-1 reporter is driven by the HIV-1 core promoter (-78 to +60), which contains all elements required for basal transcription and Tat-mediated activation. Thus, both reporters retain the native sequence context of the regulatory elements under study.

These reporter assays served as initial screening and hypothesis-testing tools. The key functional insights from reporters were then validated in more physiological contexts:

(i) The position-dependent effect of UBP1 binding (activation when upstream of TSS, repression when overlapping TSS) observed in chimeric reporters (Fig 6A-C) was independently confirmed by motif enrichment analysis of endogenous genes from RNA-seq data (Fig 6D-E).

(ii) Our immobilized template transcription assays directly measure nascent RNA production from the HIV-1 promoter, showing that MED16 overexpression reduces transcriptional output (Fig 7C). Thus, while reporter assays alone would be insufficient, our conclusion on UBP1-MED16 function are further corroborated by both global genomic analysis and direct in vitro transcription measurements.

(e) The MS experiments show promise, but the data were not shown, so it's hard to judge. The reader cannot compare/contrast the experiments, and we have no indication of the statistical confidence of the proteins identified. How many biological replicate MS experiments were performed?

The IP-MS experiments were performed with three biological replicates for each of the four antibodies against MED1, MED12, CDK8, and MED16 respectively. The complete datasets are now provided in the revised manuscript (Supplementary Table 1).

In addition, we have reanalyzed the data using a more quantitative approach. In the original submission, we classified proteins by a binary cutoff—detected or not detected in MED16 relative to other baits. In the revised analysis, we calculated a bait-preferential score for each protein by comparing its normalized abundance in the MED16 IP to its mean abundance across the other three Mediator baits. This score reflects how strongly a protein is enriched by MED16, rather than simply whether it meets an arbitrary detection threshold. A detailed description of this method is provided in the revised Methods section.

Notably, using this quantitative method, UBP1 and TFCP2 can also be detected at lower levels in some replicates of the other Mediator baits. This is consistent with our functional model: UBP1 and TFCP2 associate with the whole Mediator complex through MED16. The MED16-biased enrichment pattern, now supported by the bait-preferential score, reinforces the conclusion that MED16 serves as the primary anchor for UBP1/TFCP2 to associate with the Mediator complex.

(3) The data are over-interpreted, and alternative (and more plausible) hypotheses are ignored. Many examples of this, some of which are alluded to in the points above. For example, Med16 loss or overexpression will cause compensatory responses in cells. An expected result is that Mediator composition will be disrupted, since Med16 directly interacts with several other subunits. Also in yeast, the Robert, Gross, and Morse labs showed that loss of Med16/Sin4 causes loss of other tail module subunits, and this would be expected to cause major changes in the transcriptome. The authors also mention that yeast Med16/Sin4 "alters chromatin accessibility globally" and this would be expected to cause major changes in the transcriptome, leading to unintended consequences that will make data analysis and identification of direct Med16 effects impossible. The unintended consequences will be magnified with prolonged disruption of MED16 levels in cells (e.g., longer than 4h). These unintended consequences are hard to predict or define, and are likely to be widespread given the pivotal role of Mediator in gene expression. One unintended consequence appears to be loss of pol2 upon Med16 over-expression, as suggested by the western blot in Figure 8B. I point this out as just one example of the caveats/pitfalls associated with long-term knockdowns or over-expression.

We thank the reviewer for raising this important point. Indeed, our new Co-IP and IP-MS data confirm that MED16 knockout leads to partial dissociation of MED23, MED24, and MED25 from the Mediator complex. We agree that this structural perturbation must be carefully considered when interpreting phenotypes from MED16 depletion experiments, and we acknowledge that long-term knockdown or knockout approaches are inherently susceptible to secondary effects and cellular adaptation.

MED16 knockout impacts on the Mediator complex

However, our key conclusions are not based solely on long-term depletion experiments. Several lines of evidence, including experiments performed on short timescales without prolonged MED16 loss, converge on the same conclusion:

First, the MED16 mutant lacking its C-terminal αβ-domain cleanly separates the two functions. This mutant retains the intact N-terminal WDR domain of MED16 and can integrate into the Mediator complex, preserving its structural role and preventing tail module dissociation. Yet it completely fails to exert the competitive inhibition (squelching) effect on the UBP1 reporter that is observed with wild-type MED16 overexpression (Fig 3G), and similarly fails to inhibit HIV-1 reporter activity (Fig 5G). The sole molecular defect of this mutant is its inability to bind UBP1 (Fig 2B). Critically, this overexpression experiment is performed in a wild-type background where the endogenous Mediator complex remains structurally intact. The only variable is whether the overexpressed MED16 can interact with UBP1. This provides strong evidence that the transcriptional effects are mediated specifically through the MED16-UBP1 interaction interface, independent of MED16's structural role in maintaining Mediator integrity.

Second, the transcriptional changes we observe are not random but show specific functional coherence. If the transcriptomic changes in MED16 knockdown cells were merely secondary consequences of Mediator tail module disruption, we would expect broad, non-specific dysregulation. Instead, the UBP1-MED16 co-regulated genes are enriched for specific biological processes—angiogenesis—which largely align with the known biological functions of UBP1 (Fig 4D-E). This functional specificity is difficult to reconcile with the global, non-specific transcriptional disruption that would be expected from the total Mediator dysfunction.

Finally, we wish to address the reviewer's observation regarding Pol II levels in Figure 8B. In preparing the revised manuscript, we carefully re-examined our original Western blot data and found that the reduction of Pol II upon MED16 overexpression was not consistently reproducible across independent experiments. We have therefore replaced new Pol II blot in Figure 8B. We note that this revision does not affect any of the conclusions drawn from Figure 8, which focus on the inhibitory effect of MED16 overexpression on HIV-1 latency reversal in J-Lat cells.

We are grateful for the reviewer’s careful assessment, which encouraged us to refine our data presentation and enhance experimental rigor. The manuscript has been significantly improved as a result.

Reviewer #3 (Public review):

Weaknesses:

(1) The decrease in 1,6-hexanediol-treated cells of MED16 is modest, variable, not quantified, and internally inconsistent. For example, in Figure 1A, 1,6-hexanediol treatment should not have an impact on the level of the protein being directly IP. For MED12 (and CDK8 and MED1 to a lesser extent), 1,6-hexanediol treatment alters the level of the target protein in the IP. Along these lines, Figure 1A shows a no 1,6H-D dependent decrease in MED1 or MED12 levels in the CDK8 IP, whereas Figure 1B does show a decrease. Figure 1A shows no 1,6H-D dependent decrease in CDK8 levels in the MED1 IP, whereas Figure 1B shows a dramatic decrease. MED24 levels in the MED12 IP increase upon 1,6H-D in Figure 1A, but decrease in Figure 1B. Internal inconsistencies of this nature persist in the other Figures.

We thank the reviewer for this careful observation, and we here clarify the rationale behind our experimental design to resolve the apparent discrepancies between Figures 1A and 1B.

We first performed co-immunoprecipitation (co-IP) assays using HEK293T whole-cell lysates with three distinct antibodies targeting CDK8, MED1, and MED12, respectively, to pull down intact Mediator complexes. This setup was designed to assess whether 1,6-hexanediol (1,6-HD) impairs Mediator complex integrity and, if so, whether this perturbation affects individual subunits unequally. Across all three independent IP replicates, we observed a uniform trend: most Mediator subunits displayed only mild signal attenuation, whereas MED16 exhibited a striking, dose-dependent reduction upon escalating 1,6-HD concentrations (Fig 1A). Quantitative densitometry of these immunoblot signals has now been incorporated into the revised Figures 1A and 1B to better visualize this difference. Thus, this initial result in Fig 1A indicated that MED16’s interaction with the core Mediator complex is uniquely vulnerable to 1,6-HD disruption, irrespective of which Mediator subunit was used for immunoprecipitation.

To validate the reproducibility of this phenotype, we replicated the assay in an alternative biochemical system using HeLa nuclear extracts (NEs) (Fig 1B), with 2,5-hexanediol (2,5-HD) included as a critical negative control—this structural isomer cannot disrupt weak hydrophobic protein-protein contacts. Notably, the HeLa NE experiments in Fig 1B were carried out under elevated salt conditions (~300 mM NaCl), in contrast to the 150 mM NaCl buffer used for the HEK293T lysate assays in Figure 1A.

We want to emphasize that despite these substantial technical disparities between the two experimental setups, both datasets converge on the identical core conclusion: MED16’s association with the core Mediator complex is selectively disrupted by 1,6-HD. This consistent trend is further strengthened by quantitative analysis of immunoblot band intensities from biological replicates, which we have added to the revised figures.

Minor variability observed for other Mediator subunits between Fig 1A and 1B arises from intrinsic differences in cell type, extraction protocols, and buffer salt concentrations. For this reason, we did not draw mechanistic interpretations regarding the variable signal changes of other subunits across distinct experimental conditions; instead, our analysis focused on the robust, consistent loss of MED16 following 1,6-HD exposure.

(2) Undermining the value of Figure 1E/F, UBP1 and TFCP2 may also associate with the small amount of MED16 in the 2MDa fractions. This is not tested, and therefore, the conclusion that they just associate with the dissociable form of MED16 is not supported.

We thank the reviewer for this insightful observation. Indeed, UBP1 and TFCP2 can associate with MED16 in the context of the intact Mediator complex. Figure 1B directly supports this: under 2,5-HD conditions, where MED16 remains integrated into the Mediator, both UBP1 and TFCP2 are co-immunoprecipitated with the Mediator complex (via anti-MED1). This association is lost under 1,6-hexanediol conditions, where MED16 dissociates—suggesting that UBP1 and TFCP2 are tethered to the Mediator through MED16.

The gel filtration experiment (Fig 1D) further reveals that a portion of MED16 resolves away from the core Mediator complex under these in vitro conditions, and Figure 1E-F shows that UBP1 and TFCP2 co-immunoprecipitated with MED16 in these lower molecular weight fractions. Taken together, these data indicate that UBP1 and TFCP2 can be found in association with MED16 in two biochemically distinguishable states: one integrated into the Mediator complex, and another that has dissociated from it under our experimental conditions. We did not state that UBP1 and TFCP2 associate exclusively with the dissociated form; rather, the dissociated fraction is where we initially identified this interaction, which then led us to investigate its functional significance. The whole manuscript has been extensively rewritten now, and hopefully the confusion has been clarified.

Our functional data further demonstrate that the transcriptional activities of UBP1 and TFCP2 require the intact Mediator complex. MED16 bridges this interaction through its C-terminal αβ-domain, which binds UBP1/TFCP2, and its N-terminal WDR domain, which anchors MED16 into the Mediator. We have revised the manuscript to clearly distinguish the biochemical observation from the functional model, and we now explicitly state that UBP1 and TFCP2 associate with the Mediator complex through MED16.

(3) Domain mapping studies in Figure 2 are overinterpreted. Since the interactions could be indirect, it is not accurate to conclude "Therefore, the N-terminal WDR domain of MED16 is crucial for its integration into the Mediator complex, while the C-terminal αβ-domain is essential for interacting with UBP1-TFCP2."

We fully acknowledge that co-immunoprecipitation (co-IP) assays alone cannot discriminate direct physical binding from indirect protein associations mediated by intermediate partners. Nevertheless, our conclusion that MED16 directly interacts with UBP1 is supported by two orthogonal, independent experimental datasets.

First, GST pull-down assays using purified recombinant proteins expressed in E. coli confirm direct binding (Figure 2C). This cell-free biochemical system removes all potential cellular bridging factors, unambiguously demonstrating a physical contact between the C-terminal αβ-domain of MED16 and UBP1.

Second, our mutational functional analyses further rule out an indirect interaction model. Deletion of the unique 36-residue USP peptide in UBP1 completely abrogates MED16 binding (Figure 2F, comparison between ΔSAM and DBD truncations). Reciprocally, MED16 mutants lacking the C-terminal αβ-domain lose all capacity to associate with UBP1. Importantly, both truncation mutations yield identical functional defects in UBP1-driven and HIV-1 luciferase reporter assays (Figures 3G and 5G). If this complex assembly relied on an uncharacterized bridging protein, it would be statistically improbable for two distinct, structurally unrelated deletions (USP removal on the UBP1 side and αβ-domain truncation on the MED16 side) to independently disrupt the same indirect linkage and generate identical loss-of-function phenotypes.

Collectively, the congruent results from cell-free in vitro biochemistry and targeted mutational functional assays provide robust evidence for a direct, specific physical interaction between MED16 and UBP1.

(4) A close examination of Figure 2C undermines confidence in the association studies. The bait protein in lanes 5-8 should be equal. Also, there is significant binding of GST to UBP1 and TFCP2, in roughly the same patterns as they bind to GST-MED16 αβ. The absence of input samples makes the results even more difficult to interpret.

We acknowledge the technical limitations the reviewer pointed out. Regarding the concern about unequal protein loading across lanes 5-8, we note that lanes 7 and 8 in fact contain more GST-M16αβ bait protein than lanes 5 and 6, yet TFCP2L1 and YY1 still show no appreciable binding above the GST-alone background. This loading asymmetry therefore strengthens the conclusion that these proteins do not directly interact with MED16. For UBP1 and TFCP2, we have now provided quantification of the Western blot signals, which confirms a clear difference in binding to GST-M16αβ versus GST alone. Furthermore, these GST pull-down experiments were performed under stringent conditions—300 mM NaCl and 0.5% Tween-20 in the wash buffer—and the UBP1 and TFCP2 interactions with GST-M16αβ persisted under these conditions (lane 5-8), whereas binding to GST alone was minimal (lane 1-4). These technical details support the specificity of the observed interactions.

(5) The domain deletion mutants are utilized throughout the manuscript as evidence of the importance of the UBP1-MED16 interaction. However, in Figure 2F lanes 7 and 8, the delta-S mutant binds MED16 as well as full-length UBP1. This undermines much of the subsequent data and conclusions about specificity.

We believe the reviewer may have a misunderstanding about the experimental design. The delta-S (ΔSAM) mutant deletes the SAM motif, which is required for UBP1-TFCP2 heterodimerization, but retains the UBP1-specific sequence (USP) that mediates MED16 binding. The observation that ΔSAM still binds MED16 at levels comparable to full-length UBP1 is therefore not a discrepancy—it is the expected result and confirms that USP is the region responsible for MED16 interaction. The DBD mutant, which further removes USP, completely loses MED16 binding. Together, these deletion constructs delineate two separable interaction surfaces on UBP1: USP for MED16 binding and the SAM motif for TFCP2 heterodimerization (see Figure 2 G). The functional analysis of these domains is reflected in Figure 3B: ΔSAM, which retains MED16 binding but loses TFCP2 dimerization, shows approximately half the transcriptional activity of full-length UBP1, suggesting that both interactions contribute to full UBP1 function. To avoid similar confusion, we have renamed ΔS to ΔSAM，UBP1-SP to USP in the revised manuscript.

(6) Even if the delta-S mutant were defective for MED16 binding, the result in Figure 3B does not "confirm that MED16 is required for the transcriptional activity of UBP1,". Removal of that domain may have other effects.

We wish to first clarify a factual point: the delta-S mutant (ΔSAM) is not defective for MED16 binding. ΔSAM deletes the SAM motif that is required for UBP1-TFCP2 heterodimerization, but retains the UBP1-specific sequence (USP) that mediates MED16 binding. As shown in Figure 2F, ΔSAM binds MED16 at levels comparable to full-length UBP1. The mutant that loses MED16 binding is DBD, which further deletes USP.

The functional result in Figure 3B is therefore consistent with this biochemical mapping: ΔSAM retains MED16 binding but loses TFCP2 dimerization, and accordingly shows approximately half the transcriptional activity of full-length UBP1. This suggests that both the MED16 interaction (via USP) and the TFCP2 interaction (via the SAM motif) contribute to full UBP1 function.

More importantly, the conclusion that MED16 is required for UBP1 transcriptional activity does not rest on domain mutant analysis alone. It is supported by independent loss-of-function approaches: shRNA-mediated MED16 knockdown (Fig 3C) and CRISPR-mediated MED16 knockout across five independent frameshift clones (Fig 3E-F), both of which consistently reduce UBP1 reporter activity. The domain deletion mutants complement these approaches by mapping the specific interaction interfaces.

(7) As Mediator is critical for the activation of many genes, it is not accurate to assume that the impact of its deletion in Figure 3E/F demonstrates a direct requirement in UBP1-driven transcription. This could easily be an indirect effect.

MED16 knockout data must be interpreted cautiously due to potential Mediator destabilization artifacts. However, our conclusion does not rely solely on loss-of-function data; transient competitive overexpression and domain-mapping assays bypass chronic MED16 depletion entirely. The MED16Δαβ mutant retains full Mediator integration but loses UBP1 binding. Unlike wild-type MED16, this mutant fails to change UBP1 or HIV-1 reporter activity in 36–48 h transient assays performed in wild-type cells with intact endogenous Mediator. Dose titration further confirms that only UBP1-interacting MED16 competes for UBP1 binding to repress target promoters (Fig 3G). Since these experiments involve no permanent MED16 ablation, the observed transcriptional effects cannot be attributed to indirect global Mediator dysfunction, verifying specific MED16–UBP1 regulation.

(8) Without documenting the relative protein expression levels in Figure 3G/H, conclusions cannot be drawn about the titration experiments, nor the co-expression experiments. These findings are likely the result of squelching or some form of competition that is not directly related to the UBP1-mediated transcription. A great deal of validation would be required in order to support the model that these effects are a result of MED16 overexpression sequestering UBP1 away from holo-Mediator.

We thank the reviewer for this insightful comment. In fact, the squelching interpretation that the reviewer proposes is fully consistent with our functional model and helps reinforce a key conclusion of our study.

The reviewer correctly points out that overexpressed MED16 may compete with the endogenous Mediator complex for UBP1 binding, sequestering UBP1 away from the holo-Mediator and thereby inhibiting UBP1-driven transcription. This interpretation directly implies that UBP1 requires association with the intact Mediator complex—not merely free MED16—to activate transcription. This is precisely the model we advocate, and the squelching phenomenon observed in Figure 3G thus provides independent functional evidence for it.

The specificity of this effect is confirmed by the MED16 Δαβ mutant: the Δαβ mutant integrates into the Mediator complex normally but cannot bind UBP1, and consequently fails to exert any competitive inhibition on the UBP1 reporter (Fig 3G). This demonstrates that the squelching effect is specifically dependent on the MED16-UBP1 interaction interface. We have now documented the relative protein expression levels for these experiments in the revised manuscript.

(9) The lack of any documentation of expression levels for the various ectopic proteins in the majority of Figures, renders mechanistic claims meaningless (Figures 3, 4, 5, 6, 7, S2, S3). This is particularly relevant since the model presented for many of the results invokes concentration-dependent competition.

In the revised manuscript, we have now included Western blot data documenting the expression levels of ectopic proteins for the key experiments. The expression level data confirm that the proteins were expressed at the expected levels across the different conditions.

Recommendations for the authors:

Reviewing Editor Comments:

The individual reviews are included above. Please also note that additional instructions regarding the publication of the preprint are below.

If you choose to publish your preprint with eLife, these reviews will also be published with the article. As noted in the instructions following the reviews, you do have a chance to include a provisional response to the reviewers' comments in the provisional publication of the preprint. You will also have an opportunity to submit a revised version of the manuscript in the future. Finally, you do have the option of formally withdrawing the article from eLife after the Reviewed Preprint has been published and pursuing publication elsewhere.

Reviewer #1 (Recommendations for the authors):

(1) The writing and presentation are generally good, except that some typos and mistakes in conceptual statements (e.g., lines 407-408 "WDR-deleted MED16, which cannot interact with UBP1") often confuse the readers.

We thank the reviewer for the careful reading. We have tried out best to correct the typos and misstatements, including the error in line 407-408 where "WDR-deleted MED16, which cannot interact with UBP1" should refer to the αβ-domain deletion rather than the WDR deletion. The whole manuscript has been extensively rewritten for better clarity and preciseness.

(2) Inconsistency in the citation format between numbering and author names should be avoided.

The citation format has now been unified throughout the manuscript.

(3) Two important scientific concepts based on the published literature should be mentioned:

(a) "The position-dependent function of transcription factors (TFs) relative to the TSS" (lines 427-428) is a well-known fact in the transcription field that partly explains gene activation versus gene repression; and

(b) the statement "JQ1 that is known to reverse HIV-1 latency by antagonizing BRD4's suppression of the Tat-P-TEFb/SEC (Super Elongation Complex) interaction" (lines 384-386) is only partially correct and is a property unique to the BRD4 long (BRD4-L) isoform.

We thank the reviewer for this suggestion. We have now cited the relevant literature on the position-dependent function of transcription factors relative to the TSS, including the recent study by Duttke et al. (3), which systematically demonstrated that the effect of TF binding on transcription initiation is highly position-dependent. Our findings provide a clear example that UBP1 activates transcription when its binding motif is located upstream of the TSS but represses when the motif overlaps the TSS, which exemplifying this concept.

We have corrected the statement to specify that JQ1 antagonizes the BRD4 long isoform (BRD4-L)-mediated suppression of Tat-P-TEFb/SEC, and have noted that the BRD4 short isoform (BRD4-S) may also contribute to HIV-1 promoter repression and JQ1-mediated latency reversal.

(4) Another relevant concept is that BRD4 short (BRD4-S) isoform-inhibited HIV-1 promoter activity could be potentially alleviated by JQ1, leading to latency reversal. Besides the aforementioned points and those comments already provided in the Public Review, the involvement of TFCP2 and YY1 in MED16-UBP1 function has not been clearly established and will require more careful and detailed investigation.

We thank the reviewer for these suggestions. Regarding BRD4-S, we have added a discussion of the BRD4 short isoform's role in HIV-1 promoter regulation and its potential contribution to JQ1-mediated latency reversal.

Regarding the involvement of TFCP2 and YY1, we agree that their precise roles within the MED16-UBP1 regulatory axis have not been fully dissected in the current study. Our data indicate that TFCP2 can dimerize with UBP1 to enhance transcriptional activation (Fig 3H), and YY1 is present in the MED16-containing fractions (Fig 1E), consistent with its previously reported role in HIV-1 silencing. However, the molecular details of how TFCP2 and YY1 contribute to MED16-UBP1 function—and whether they act in the same or parallel pathways—remain to be determined. We have acknowledged this as a limitation and a direction for future investigation in the revised Discussion.

Reviewer #2 (Recommendations for the authors):

(1) The data from Figure 1F are hard to interpret, the blot quality is poor for Med23, 16

We acknowledge that the blot quality for MED23 and MED16 in Figure 1F was suboptimal in the original submission. We have now provided better Western blot, which confirms the specific co-immunoprecipitation of MED16 and MED23 with TFCP2 in both HeLa nuclear extract and the gel filtration fractions. Importantly, the reciprocal co-IP (anti-MED16 IP blotted for TFCP2 and UBP1, Fig 1E) independently validates this interaction.

(2) Figure 2: No MW markers on gels. Also, these types of experiments are ok for pilot studies, but they aren't rigorous. Only a few proteins were tested, and it's easy to get false positives with how the experiments were completed. Finally, the Mediator proteins were tested as single isolated proteins rather than in their native context. As isolated proteins, they likely have large exposed hydrophobic patches (normally bound to other subunits), and this can generate non-specific "sticky" interactions that can lead to false positive results.

We have now included MW markers on the gel images in the revised Figure 2.

The reviewer raises a valid general concern about domain mapping experiments. However, we wish to emphasize that the readout in Figure 2B and 2F is the binding of exogenously expressed truncated proteins to their endogenous interaction partners—MED1, TFCP2, MED16, and MED23—which are natively folded and assembled in their physiological environment. This is not a situation where two isolated proteins are mixed in a test tube. Furthermore, the deletion series themselves contain important internal controls against non-specific "stickiness": In Figure 2B, progressive deletion of the WDR domains leads to a gradual reduction in MED1 binding, consistent with the predicted structural role of the β-propeller (WDR domain). The truncation at the αβ-domain specifically abolishes TFCP2 binding but does not affect its association with Mediator complex as indicated by MED1. If the truncated proteins were non-specifically sticky due to exposed hydrophobic patches, one would expect uniform binding across all truncations or random binding patterns, which is not observed. And in Figure 2F, deletion of the 36-amino-acid USP abolishes its MED16 binding, while deletion of the SAM motif does not. This sequence-specific differential effect is inconsistent with non-specific hydrophobic interactions.

The domain mapping experiments in Figure 2 were not designed as an unbiased screen but as hypothesis-driven tests guided by the IP-MS results identifying UBP1/TFCP2 as MED16 interactors, sequence and structural analysis predicting the roles of the WDR and αβ-domains. The conclusions from these experiments are corroborated by subsequent functional assays: the USP-deleted mutant and the MED16 Δαβ mutant show corresponding functional defects in both UBP1 reporter and HIV-1 reporter assays (Fig 3G, Fig 5G), providing strong support that the mapped interaction interfaces are functionally relevant.

(3) Figure 3: UBP1 reporter is artificial, 6X UBP1 sites upstream of a TATA box.

We appreciate this comment and wish to clarify a misunderstanding. The UBP1 reporter used in Figure 3 is not an artificial construct assembled from random elements. Its backbone is derived from the mouse α-globin gene promoter, a well-established natural target of UBP1/TFCP2, which contains two endogenous UBP1/TFCP2 binding sites. In order to increase the reporter's sensitivity to UBP1/TFCP2, we expanded the original two sites to six. Critically, this expansion did not alter the native architecture of the promoter—the relative positions of the transcription factor binding sites, the CCAAT box, the TATA box, and the TSS remain unchanged from the original α-globin promoter. This reporter therefore retains the physiological spacing and topological organization of a genuine UBP1 target promoter, and its low basal activity and high inducibility by UBP1 make it a well-suited tool for modeling UBP1 transcriptional activity. We have further clarified this description in the revised manuscript.

(4) Figure 3D: Med16 KD not compelling, perhaps 50% knockdown? Hard to tell because no replicates and no quantitation.

We acknowledge that the shRNA-mediated MED16 knockdown in Figure 3D was partial and that the original presentation lacked replicate quantification. We initially tested three independent shRNAs targeting MED16, of which only one achieved appreciable knockdown. Recognizing this limitation, we turned to a more definitive loss-of-function approach: CRISPR-Cas9-mediated MED16 knockout. Five independent frameshift KO clones were generated in mouse HT cells, each verified by sequencing and Western blotting (Fig 3F), and all five clones consistently showed reduced UBP1 reporter activity compared with wild-type cells (Fig 3E). The consistency between the shRNA knockdown and CRISPR KO reinforce the conclusion that MED16 is required for UBP1 transcriptional activity.

(5) Figure 3E: Med16 KO will be a different cell line because of genetic adaptation. That is, an RNA-seq experiment is expected to result in many gene expression changes compared to the parental line. This could contribute to differences, or differences could result from altered Mediator composition to compensate for Med16 loss, but this wasn't addressed by the authors. A rapid degron method would be much more rigorous.

We thank the reviewer for raising this important concern. We agree that prolonged MED16 loss can elicit cellular adaptation and compensatory changes in Mediator composition, as our own new data on tail module dissociation confirm. This makes it challenging to disentangle direct and indirect effects using constitutive KO models alone.

However, two independent lines of evidence support a direct role for MED16 in UBP1-mediated transcription that cannot be explained by adaptive responses:

First, acute shRNA-mediated knockdown produces the same phenotypic trend. shRNA operates on a shorter timescale (days) without prolonged selection for compensatory mutations, yet it also reduces UBP1 reporter activity (Fig 3C). The convergence of acute and chronic loss-of-function approaches suggests the phenotype is driven by MED16 loss itself rather than by the secondary adaptations that accumulate over time in KO cells.

Second, and most critically, the MED16Δαβ mutant experiment completely bypasses the issues of genetic adaptation and Mediator structural disruption (Fig 3G). This experiment is performed in wild-type cells with an intact endogenous Mediator complex, without any MED16 depletion and without any long-term selection. The MED16Δαβ mutant retains full ability to integrate into the Mediator complex (its WDR domain is intact) but cannot bind UBP1. Overexpression of wild-type MED16 competitively inhibits the UBP1 reporter due to the squelching effect, whereas overexpression of MED16Δαβ at comparable levels does not (Fig 3G). The sole molecular difference is that MED16Δαβ doesn`t contain the UBP1-binding interface. Because this experiment involves no MED16 loss and no adaptive period, the result directly demonstrates that MED16-UBP1 interaction is required for UBP1 transcriptional activity, entirely independent of concerns about clonal adaptation.

We appreciate the reviewer's suggestion of a rapid auxin-inducible degron approach. While acute depletion strategies offer clear advantages in minimizing cellular adaptation, they are not without limitations. First, degron-based depletion rarely achieves 100% elimination of the target protein, and residual protein may confound interpretation. Second, and particularly relevant to our system, degron-mediated degradation may preferentially target free, unassembled subunits while sparing proteins already stably integrated into large macromolecular complexes such as the Mediator—potentially creating a lag or even complete resistance to degradation for the pool of MED16 that is assembled into the complex. Given these considerations, we believe our convergent strategy—combining shRNA knockdown, CRISPR-mediated knockout in multiple independent clones, and the Δαβ mutant that bypasses structural perturbation—provides a robust and complementary set of evidence that does not rely on any single depletion method.

(6) The title for Figure 3 is "Med16 is required for UBP1 activation of gene transcription", but that's not conclusively shown by these experiments.

We agree that the original title of Figure 3 may overstate. We have revised the title to " MED16 acts through the Mediator complex to support UBP1-mediated transcriptional activation ", which may better reflect the data. As discussed in our responses above, this conclusion is supported by converging evidence from shRNA knockdown (Fig 3C), CRISPR-mediated KO across multiple clones (Fig 3E-F), and the specific loss of function observed with the MED16Δαβ mutant (Fig 3G).

(7) Figure 5: Med16 protein levels were never measured, nor were other transfected proteins. Reporter assay data are correlative only; no evidence for direct effects, and only 2 other MED subunits are probed. Again, this is a starting point, but nothing to use for concrete conclusions about Med16.

We have now provided Western blot data documenting MED16 and other transfected protein levels for the experiments in Figure 5, confirming that the proteins were expressed at the expected levels. We agree that plasmid-based reporter assays alone cannot establish direct transcriptional mechanisms. The HIV-1 reporter experiments in Figure 5 serve as an initial functional readout, and their conclusions are corroborated by more direct approaches: (i) the position-dependent specificity demonstrated by chimeric reporter assays (Fig 6A-C); (ii) the independence of the effect from TAR-mediated elongation and HDAC activity (Fig S2), ruling out alternative mechanisms; and (iii) the immobilized template transcription assays, which directly measure nascent RNA production and PIC assembly on the HIV-1 promoter (Fig 7). The reporter data should be viewed as one component of a larger body of convergent evidence rather than as standalone proof.

MED23 and MED24 were chosen as the most relevant controls because they, along with MED16, form the core of the tail module submodule that creates the transcription factor interaction platform. They are structurally adjacent to MED16 within the same submodule. The fact that neither MED23 nor MED24 overexpression affects HIV-1 reporter activity (Fig 5C), whereas MED16 overexpression does so in a UBP1-binding-dependent manner (Fig 5C, 5E), demonstrates that the inhibitory effect is specific to MED16 and not a generic consequence of overexpressing a Mediator tail subunit.

(8) Figure 6: similar concerns to Figure 5. The promoters are artificial, derived from a chicken beta-actin promoter with artificially inserted UBP1 binding sequences.

We acknowledge that the chimeric promoters in Figure 6 are indeed deliberately designed. To address this specific question: Does the binding position of the UBP1 relative to the TSS dictate whether UBP1-MED16 activates or represses transcription?

To isolate position as the sole variable, we needed a backbone promoter that (i) is stably and robustly active in the cell lines used, (ii) does not itself contain endogenous UBP1 binding sites that would confound the interpretation, and (iii) allows precise placement of the UBP1 binding motif either upstream of or overlapping the TSS. The chicken β-actin promoter meets all three criteria and is a widely used tool in molecular biology for such controlled promoter engineering experiments. Notably, its high GC content around the promoter region is also a feature shared with the HIV-1 core promoter, making it a particularly relevant backbone for the HIV-1 TSS chimeric constructs.

The key controls for this experiment are internal: in the same backbone, placing the UBP1 binding site at the TSS results in MED16-UBP1-dependent repression, whereas placing it upstream does not (Fig 6A vs 6B). Furthermore, this position-dependent effect requires the MED16-UBP1 interaction interface, as the MED16 Δαβ mutant and the UBP1 DBD mutant both fail to produce the inhibitory effect (Fig 6A). The differential outcome between the two constructs—which differ only in the placement of the UBP1 binding site—cannot be attributed to the artificial nature of the backbone, because the backbone is identical in both cases.

(9) Figure 6D: Not rigorous, no statistical analysis, cannot determine whether the modest changes are biologically relevant/meaningful. Also, data were obtained through indirect measurements and long timeframes, so any direct effects from possible UBP1 binding are probably lost.

The differentially expressed genes in Figure 6D were selected using DESeq2 with a significance threshold of p < 0.05. The top 50 upregulated and top 50 downregulated transcripts were then used for UBP1 motif enrichment analysis around their promoters. We have clarified these selection criteria in the revised figure legend.

Regarding the concern that indirect measurements over long timeframes, we now have additional evidence that UBP1 binding effects are preserved in our RNA-seq data. Motif analysis of UBP1-upregulated genes using HOMER identified the CP2 family binding motif (CNRG-N6-CNRG) as the most significantly enriched motif. This indicates that UBP1-binding-dependent transcriptional effects are not lost in our dataset.

Motif enrichment result of the UBP1 up-regulated genes

Furthermore, when these UBP1-upregulated and UBP1-downregulated genes are analyzed separately for motif position relative to the TSS, a clear pattern emerges: UBP1-upregulated genes show motif enrichment flanking the TSS, whereas UBP1-downregulated genes show enrichment overlapping the TSS (Fig 6D). If the transcriptional changes were non-specific secondary consequences of prolonged UBP1 overexpression, one would not expect this position-specific distribution. Finally, this positional specificity is corroborated by the chimeric reporter experiments (Fig 6A-C), where moving the UBP1 binding site across the TSS in an otherwise identical promoter directly switches the regulatory outcome. The concordance among the de novo motif discovery, the genome-wide positional analysis, and the functional reporter assays supports the biological relevance of the motif enrichment pattern.

(10) The title for Figure 6 is "UBP1 binding site determines the UBP1-Med16 inhibition in HIV-1 transcription" but the data do not convincingly show this.

We appreciate this suggestion. We have revised the title to "The position of the UBP1 binding site relative to the TSS dictates whether MED16-UBP1 activates or represses transcription.". For clarification, Figure 6 not only demonstrates that the UBP1 binding position dictates MED16-UBP1-mediated inhibition in the HIV-1 context, but also revealed that this positional effect could be generalized to endogenous genes: UBP1 motifs flanking the TSS are associated with activation, while motifs overlapping with the TSS are associated with repression.

(11) Figure 7: The data and the experiment do not support the claim in the title that "MED16-UBP1 complex prohibited PIC formation of HIV-1 transcription". I cannot convince myself that there are meaningful differences between the results shown in 7B, but it's hard to judge because there are no replicates and no quantitation.

We have revised the title from "MED16-UBP1 complex prohibited PIC formation of HIV-1 transcription" to "MED16 interferes with HIV-1 PIC assembly at the TSS via its interaction with UBP1." We have also provided quantification of the Western blot signals from replicate experiments in Figure 7B, which confirms statistically significant differences in CDK8 binding, Pol II pSer5 signals, and TFIIB retention between conditions. We note that the key observations in Figure 7B were consistently observed across multiple pilot experiments prior to the final dataset presented in the manuscript. Thus, the reported differences are reproducible and not based on a single experiment.

(12) The immobilized template experiments had only -78 upstream and +60 downstream of the TSS. By my understanding, much more upstream DNA is needed to allow a PIC to assemble, because the 5'-end bead attachment will block. I think this was outlined in an article by the Carey lab, but I'm not sure. A point here is that it would be reassuring for the authors to show that PIC assembly was occurring and that transcription was responsive to a transcription factor, similar to what the Carey lab has done with immobilized templates. See, for example, Figure 1 in Lin & Carey Curr Protocol Mol Biol 2012 Ch12 unit 12.14.

We thank the reviewer for raising this technical consideration. We respectfully offer a different perspective based on current structural data on the human PIC.

A recent cryo-EM study of the human Mediator-bound preinitiation complex (4) demonstrated that the PIC spans approximately 50 bp upstream and downstream of the TSS. The DNA templates used for structural determination in that study were generally within 105 bp upstream and 63 bp downstream of the TSS. Our HIV-1 template (-78 to +60, i.e., 78 bp upstream and 60 bp downstream of the TSS) falls within this well-characterized range and provides sufficient space for PIC assembly.

Regarding the choice of the -78 boundary: the HIV-1 core promoter is defined as -78 to +60, and the region from -78 to -105 encompasses the HIV-1 enhancer, which contains additional regulatory elements including NF-κB and NFAT binding sites. We deliberately excluded these enhancer elements to avoid confounding regulatory inputs and to focus specifically on the UBP1 binding site at the TSS, which is the central focus of our model. The robust luciferase mRNA production observed in wild-type nuclear extract (Fig 7C) provides direct functional evidence that PIC assembly and transcription initiation proceed efficiently on this template under our conditions.

Promoter templates used for human PIC assembly (Chen et al.

, Science 372, eaba8490, 2021): templates within −105 to +65 bp of the TSS support PIC formation

(13) Related to the above concern, the qPCR analysis from the extract experiments may be misleading because of contaminating nucleic acids in the extract. More control experiments are needed, such as no NTP controls.

We thought this concern has been addressed by our experimental design.

First, the PIC step in our experimental design (Fig 7B, "PIC") is the no-NTP control: nuclear extract is incubated with the immobilized template without NTPs. After washing, the proteins bound at this step are analyzed by Western blot, and importantly, there is no transcription occurred at this step. Transcription is then initiated by adding NTPs (Fig 7B, "Trx."), and the supernatant from this step is collected for qPCR analysis of transcription.

Second, it is unlikely that the contaminating nucleic acids from the nuclear extract could confound our qPCR results, because the qPCR analysis specifically detects Luciferase mRNA transcribed from the immobilized HIV-1-Luciferase template, which is absent from eukaryotic genomes; and the endogenous genomic DNA or RNA cannot generate a false-positive signal.

Lastly, after transcription, the immobilized templates are retained on the streptavidin beads and removed, while only the supernatant containing newly synthesized RNA is collected for qPCR. Any residual template DNA or extract-derived nucleic acids bound to the beads are therefore excluded from the qPCR sample.

(14) Some references are made with the format of (author name, year), but others use numbers.

The citation format has been unified throughout the manuscript.

(15) On line 111-115, it is stated that 2,5 HD somehow "does not interfere with weak hydrophobic interaction due to different positions of hydroxyls" with no reference. Both 1,6 HD and 2,5 HD are hydrophobic molecules, and they will disrupt protein structure. The authors make a point about the differential results with 1,6 HD vs. 2,5 HD, but it's not compelling and not justified.

We have now provided the reference to support this statement. The specificity of the 1,6-HD effect is further clarified as follows.

1,6-HD and 2,5-HD are structural isomers sharing the same chemical formula and molecular weight. The key difference lies in the positions of their hydroxyl groups (See Author response image 6). The position of the two hydroxyl groups changes both the molecule’s shape and how it interacts with water and hydrophobic surfaces.

In 1,6-HD, the two -OH groups sit at the ends of the chain, giving a more symmetric, flexible molecule, while in 2,5-hexanediol the -OH groups are closer to the middle, which changes polarity distribution and local geometry (5, 6). The distinctive structures matter because hydroxyl groups strongly hydrogen-bond with water. When they are terminal and well separated in 1,6-HD, the molecule can present a more balanced hydrophilic “cap” at both ends while keeping a relatively long hydrophobic center, which helps 1,6-HD partition into and disturb weakly interacting hydrophobic regions in condensates or hydrophobic interfaces (Author response image 6). By contrast, moving the hydroxyls inward in 2,5-HD does not help with disrupting the hydrophobic interfaces (Author response image 6). That is why 2,5-HD is often used as a comparison compound: it has the similar chemical formula, but the different hydroxyl pattern usually makes it a weaker or at least differently acting perturbant.

Differential disruption of hydrophobic interfaces by 1,6-HD vs 2,5-HD

(16) On lines 117-120, the authors claim that Med16 is "integrated into the Mediator complex through weak hydrophobic interaction but not ionic interaction" based upon a few IP results. They then state (lines 130-131) that "the interaction between Mediator and Med16 is an unstable hydrophobic interaction." These statements make no sense based on any data shown, and they ignore the fact that Med16 purifies with Mediator from many different purification protocols. It's safe to say that its interaction with Mediator results from strong, extensive hydrophobic interfaces, which is further supported by actual structural data. The authors should check their reasoning here because they are over-interpreting their results, and this will be misleading to readers.

We thank the reviewer for this important suggestion. The reviewer is right that our original phrasing—particularly "unstable hydrophobic interaction"—was imprecise and inconsistent with the established structural data. As the reviewer notes, MED16 co-purifies with the Mediator complex through multiple purification protocols, and recent cryo-EM structures (4) reveal that MED16 engages in extensive hydrophobic interfaces with neighboring tail module subunits, particularly MED24 and MED25. We fully agree that this is a strong, hydrophobic interaction in the physiological context.

Our 1,6-hexanediol experiments demonstrated that the MED16-Mediator interface does have a hydrophobic character, which renders the sensitivity to 1,6-HD disruption. This does not mean the interaction is weak. We have now carefully described the relevant statements in the revised manuscript.

Reviewer #3 (Recommendations for the authors):

(1) It is not accurate to claim that "198 (genes) were repressed by shMED16 knockdown (Fig 4B and 4C)". These genes were simply not activated by UBP1 in the knockdown cells.

We thank the reviewer for this careful distinction. We have revised the text to state that the UBP1-activated genes were “profoundly downregulated upon MED16 depletion” rather than “repressed by shMED16 knockdown”. The biological point remains unchanged: MED16 is required for UBP1 to activate this set of target genes.

(2) Like the majority of experiments presented, the RNA-seq presented in Figure 4 may reflect indirect effects of MED16 loss. As such, the results do not support the claim that "This result demonstrated that UBP1 and MED16 collaborate to regulate transcription".

We acknowledge that RNA-seq data from MED16 knockdown cells alone cannot distinguish direct from indirect effects. However, we did not rely on the RNA-seq data alone to support a collaborative role for UBP1 and MED16.

The key evidence comes from experiments comparing wild-type MED16 with the MED16Δαβ mutant. MED16Δαβ integrates into the Mediator complex normally but cannot bind UBP1 (Fig 2B). When overexpressed in wild-type cells, wild-type MED16 inhibits the UBP1-driven reporter activity through a squelching effect. In contrast, MED16Δαβ has no effect on the UBP1 reporter assay (Fig 3G), and the same is true for the HIV-1 reporter assays (Fig 5G). Thus, MED16 must both integrate into the Mediator complex and directly bind UBP1 to support UBP1-mediated transcription.

In addition, the genes identified as UBP1-MED16 co-regulated by RNA-seq are enriched for specific biological processes—angiogenesis (Fig 4D-E)—that align with the known physiological functions of UBP1. Such functional coherence would be unexpected if the transcriptional changes were merely non-specific secondary consequences of Mediator disruption.

Thus, while we agree that RNA-seq alone has limitations, the convergent evidence from the MED16Δαβ mutant and the functional specificity of the co-regulated gene set supports a collaborative role for UBP1 and MED16 in transcriptional regulation.

(3) The data in Figure 4 are not presented with controls documenting the relative expression levels of MED16 and UBP1. For example, it would be important to understand the level of UBP1 overexpression achieved, relative to endogenous UBP1 levels in HeLa cells.

We agree with the reviewer that documenting expression levels is important. Author response image 7 are the CPM values for UBP1 from our RNA-seq data, confirming robust overexpression in the UBP1-overexpression condition compared with the control. The MED16 protein levels in the knockdown cells can be found in Fig 3D, as the same MED16-knockdown HeLa cell line was used for the RNA-seq experiment.

UBP1 expression level derived from RNA-seq

Regarding the physiological relevance of UBP1 overexpression, several observations support that the transcriptional effects we observe are specific rather than artifacts of overexpression: (i) the UBP1-activated genes are enriched for pathways known to be regulated by UBP1—angiogenesis (Fig 4D-E)—rather than showing random, non-specific activation; (ii) the activation of these genes is attenuated upon MED16 knockdown, demonstrating dependence on the MED16-UBP1 interaction; and (iii) in reporter assays, the MED16 mutant lacking its C-terminal αβ-domain fails to competitively inhibit UBP1 driven reporter transcription even when UBP1 is similarly overexpressed (Fig 3G), indicating that the competitive inhibition of the UBP1 reporter depends specifically on the MED16-UBP1 interaction interface, rather than on the amount of UBP1 protein overexpressed.

(4) The enhancer-promoter architecture of the luciferase construct used in Figure 3 is not sufficiently described. Are the elements far enough apart that Mediator-dependent looping is required for transcription, for example?

We appreciate the reviewer's question. The UBP1 reporter construct used in Figure 3 is derived from the mouse α-globin gene promoter, a well-established natural target of UBP1/TFCP2(7, 8). The native α-globin promoter contains two endogenous UBP1/TFCP2 binding sites; to increase the reporter's sensitivity to UBP1, we expanded these to six tandem copies. Critically, this expansion preserved the native architecture of the promoter—the relative positions of the UBP1 binding sites, the CCAAT box, the TATA box, and the TSS were not altered. The entire reporter construct spans only 198 bp in total length. At this scale, all regulatory elements reside within a compact proximal promoter region, with no element separated by a distance that would require DNA looping for functional communication. The reporter therefore measures UBP1-dependent transcriptional activation from the proximal promoter, consistent with the canonical role of Mediator in facilitating PIC assembly rather than mediating long-range enhancer-promoter interactions. We have added a detailed description of the construct architecture to the Methods section.

(5) The identification of matches to the UBP1 consensus sequence in a promoter does not prove that this is how UBP1 overexpression is impacting their transcription. "highly matched UBP1 motifs were identified in their promoters (Fig. 4F), leading to their activation when UBP1 was overexpressed."

We agree with the reviewer that the presence of a UBP1 consensus motif in a promoter does not by itself prove direct regulation. We have revised the original statement: the activation of these genes by UBP1 overexpression is now described as "associated with" the presence of UBP1 motifs in their promoters.

We acknowledge that the ideal experiment to demonstrate direct UBP1 occupancy would be ChIP-seq. However, ChIP-grade antibodies against UBP1 are not commercially available, and no UBP1 ChIP-seq datasets exist in public databases. We attempted to carry out the ChIP-seq ourselves but were unable to obtain sufficiently good-quality data.

In the absence of ChIP data, three lines of evidence indicate that the identified UBP1 motifs are functionally relevant. First, motif analysis of UBP1-upregulated genes using HOMER identified the CP2 family binding motif (CNRG-N6-CNRG) as the most significantly enriched motif. This demonstrates that UBP1-binding-dependent transcriptional could be preserved in our RNA-seq data. Second, when we examine the positional distribution of these motifs, a clear pattern emerges: UBP1 motifs flank the TSS in activated genes, but overlap the TSS in repressed genes (Fig 6D). Indirect effects would not produce this position-specific pattern. Third, our chimeric reporter experiments (Fig 6A-C) directly test this position effect: moving the UBP1 binding site upstream of the TSS leads to activation, while moving it to overlap the TSS leads to repression. Together, these three independent approaches support the conclusion that UBP1 regulates these genes through the motifs we identified. See Author response image 4.

(6) Lines 266 and 277 are redundant.

We have carefully reviewed lines 266 and 277. Line 266 is the topic sentence (caption) of the section ("MED16 collaborates with UBP1 to inhibit HIV-1 transcription"), which introduces the overall conclusion, while line 277 describes the specific experimental result ("MED16 inhibited HIV-luciferase activity in a dose-dependent manner"). These serve distinct rhetorical functions. We have ensured that no other redundant phrasing exists elsewhere in the section.

(7) The mere correlation of binding site position with activation status does not prove "that UBP1 may act as either a transcriptional activator or repressor, depending on distance of its binding site to TSS."

We appreciate the reviewer's concern regarding the distinction between correlation and causation. The genome-wide analysis (Fig 6D) reveals an association between UBP1 motif position and transcriptional outcome, which prompts us to postulate that the UBP1 binding position could dictate the transcriptional activity level. The causal evidence comes from the chimeric reporter experiments (Fig 6A-C), in which we directly manipulated the position of the UBP1 binding site relative to the TSS in an otherwise identical promoter background. Placing the UBP1 binding site at the TSS resulted in MED16-UBP1-dependent repression, whereas placing it upstream of the TSS did not. This controlled experiment demonstrates that changing only the position of the UBP1 binding site is sufficient to switch the regulatory outcome, thereby establishing a causal relationship between binding site position and transcriptional effect. We have revised the text to clearly distinguish the correlational genome-wide analysis from the causal reporter experiments.

(8) In Figure 7, MED16 overexpression results in less CDK8 binding. This undermines the claim that MED16 overexpression lowers transcription by binding to UBP1, and suggests an indirect effect.

We appreciate the reviewer's careful observation. The reduction in CDK8 binding upon MED16 overexpression is consistent with our model: MED16, through UBP1 occupying the TSS, interferes with PIC assembly, of which CDK8 is a component. Importantly, the input levels of CDK8 and other Mediator subunits remained unchanged across the three conditions (Fig 7B), indicating that the reduced CDK8 signal in the PIC is not due to altered protein expression.

In support of the specificity of the transcriptional effect, MED16Δαβ—which integrates into the Mediator complex but cannot bind UBP1 (Fig 2B)—fails to competitively inhibit either the UBP1 reporter (Fig 3G) or the HIV-1 reporter (Fig 5G), confirming that the transcriptional outcome depends on the MED16-UBP1 interaction.

(9) Figure 8F is not referenced in the text or legend.

Figure 8F has now been referenced in the Results and Discussion section.

Reference:

(1) Wang G, Cantin GT, Stevens JL, Berk AJ. Characterization of mediator complexes from HeLa cell nuclear extract. Mol Cell Biol. 2001;21(14):4604-13.

(2) Yamazaki T, Souquere S, Chujo T, Kobelke S, Chong YS, Fox AH, et al. Functional Domains of NEAT1 Architectural lncRNA Induce Paraspeckle Assembly through Phase Separation. Molecular Cell. 2018;70(6):1038-+.

(3) Duttke SH, Guzman C, Chang M, Delos Santos NP, McDonald BR, Xie J, et al. Position-dependent function of human sequence-specific transcription factors. Nature. 2024;631(8022):891-8.

(4) Chen X, Yin X, Li J, Wu Z, Qi Y, Wang X, et al. Structures of the human Mediator and Mediator-bound preinitiation complex. Science. 2021;372(6546).

(5) Zheng T, Wake N, Weng SL, Perdikari TM, Murthy AC, Mittal J, et al. Molecular insights into the effect of 1,6-hexanediol on FUS phase separation. EMBO J. 2025;44(10):2725-40.

(6) Romero CM, Páez MS, Miranda JA, Hernández DJ, Oviedo LE. Effect of temperature on the surface tension of diluted aqueous solutions of 1,2-hexanediol, 1,5-hexanediol, 1,6-hexanediol and 2,5-hexanediol. Fluid Phase Equilibr. 2007;258(1):67-72.

(7) Lim LC, Swendeman SL, Sheffery M. Molecular cloning of the alpha-globin transcription factor CP2. Mol Cell Biol. 1992;12(2):828-35.

(8) Chae JH, Kim CG. CP2 binding to the promoter is essential for the enhanced transcription of globin genes in erythroid cells. Mol Cells. 2003;15(1):40-7.

Version 2

Abstract

The Mediator complex acts as a master transcriptional co-regulator, yet the mechanisms by which its individual subunits mediate gene-specific transcription remain incompletely defined. Here, we report that the Mediator subunit MED16 physically associates with the UBP1-TFCP2 heterodimer, which controls the context-dependent transcriptional activation or repression. Biochemical purification combined with mass spectrometry validated MED16 as a bona fide interactor of UBP1-TFCP2. Integrative transcriptomic analyses demonstrated that UBP1 recruits the Mediator complex through MED16 to activate a panel of silenced genes associated with lung homeostasis, angiogenesis and cell proliferation. By contrast, MED16-UBP1 co-occupancy at the HIV-1 transcription start site (TSS) represses viral transcription and sustains HIV latency by compromising pre-initiation complex assembly. Domain-mapping experiments revealed that MED16 must integrate into the intact Mediator complex to exert either activating or repressive function. Genomic analysis uncovered a positional regulatory logic: UBP1-TFCP2 binding motifs adjacent to the TSS promote transcription, while those overlapping the TSS elicit repression. Collectively, our work identifies an MED16-UBP1 regulatory module with dual transcriptional activity and highlights this signaling axis as a potential target for HIV-1 interventions.

Introduction

Transcription by RNA polymerase II (Pol II) requires the assembly of a pre-initiation complex (PIC) at gene promoters. Among the co-activators that facilitate this process, the multi-subunit Mediator complex plays a central role (1–8). The Mediator complex is composed of ∼30 subunits in mammals and structurally divided into four modules: Head, Middle, Tail, and Kinase. The scaffold subunit MED14 joins the Tail, Middle, and Head modules together to form the “core Mediator”. The tail module is composed of MED15, MED16, MED23, MED24, MED25, MED27, MED28, MED29, and MED30; and the main function of the tail module is to provide an interactional platform for transcription factors to recruit Mediator complex to gene promoters and enhancers for further regulation of gene expression (8–10). Among all tail subunits, MED16, MED23, MED24, and MED25 form a submodule that accounts for ∼70% molecular weight of the tail module and creates a large platform for interactions of transcription factors (11). Previous studies have found MED23 interacts with multiple transcriptional factors and controls diverse gene expression programs at multiple stages of transcription processes (12–15). It prompts us to question whether other subunit in this submodule play similar or distinctive roles in transcriptional regulation, especially MED16 that is structurally conserved from yeast to mammals and acts as a critical regulator in yeast and plants. Sin4 (yeast MED16) plays both positive and negative roles in the transcriptional regulation of many genes in yeast, and loss of Sin4 alters chromatin accessibility globally (16). In plant cells, MED16 controls endoreduplication and cell growth by repressing the expression of APC/C activator genes CCS52A1/A2 (17–21). MED16 also controls plant cold response, chemical and wound response, and iron homeostasis at transcriptional levels (17–21). However, the functions of MED16 in mammalian cells are poorly understood. Among a few reports, MED16 was reported as a key factor in NRF2-activted gene expression via its direct interaction with NRF2. Knockout of MED16 reduces NRF2-dependent anti-oxidative gene expression (21, 22). Additionally, MED16 was down-regulated in papillary thyroid cancer, leading to increased TGF-β signaling and radioiodine resistance (23). Though the exact molecular mechanisms by which MED16 regulates gene expression remain to be further elucidated, and its roles in animal development and diseases are barely investigated.

UBP1 (as known as LBP-1) and TFCP2 (as known as LSF) belong to a subfamily of the TFCP2/Grainyhead transcription factors family. The amino acid sequences of TFCP2 and UBP1 share 88% identity and they can form both homo-and heterodimers (24–26). UBP1 and TFCP2 bind to the consensus sequence CNRG-N6-CNR(G/C) (27). They function as a transcription activator or repressor to regulate a number of viral and cellular genes, including mouse α-Globin, MMP-9, CYP11A1, IL-4, SV-40, and HIV-1 (28–32). UBP1 and TFCP2 are involved in various biological processes, including embryonic development, angiogenesis, tumorigenesis, and HIV-1 latency (28, 29, 31–38). There is a canonical UBP1 binding motif within the HIV-1 core promoter which overlaps with the transcription start site (TSS). UBP1 has been implicated in HIV-1 transcriptional repression through several reported mechanisms, including elongation restriction and recruitment of YY1/HDAC1 (32–34). However, how UBP1 and TFCP2 regulate cellular gene expression, and what are the genome-wide target genes of UBP1 and TFCP2 still remain unclear.

In this study we identified an interaction between MED16 and UBP1-TFCP2 dimer utilizing gel-filtration chromatography and immunoprecipitation-mass spectrometry (IP-MS). Domain mapping revealed that MED16 uses its N-terminal WDR domain to anchor within the Mediator complex and its C-terminal αβ-domain to bind UBP1. Functionally, MED16 must act through the Mediator complex to support UBP1-mediated transcriptional activation, including the activation of pulmonary surface molecule genes. Conversely, MED16 cooperates with UBP1 within the Mediator complex to inhibit HIV-1 transcription at the viral transcription start site. At the genomic scale, UBP1 activates gene expression when the UBP1-TFCP2 binding motif is located near TSS, but inhibits gene expression by obstructing normal PIC formation when the UBP1-TFCP2 binding motif overlaps with TSS. Therefore, the position of the UBP1 binding site relative to the TSS determines whether UBP1-MED16 activates or represses transcription. These findings provide mechanistic insights into UBP1-mediated HIV-1 repression and suggest a potential strategy for HIV-1 therapy.

Result

A fraction of MED16 biochemically dissociates from the core Mediator complex

Multiple Mediator subunits consist of large intrinsically disordered regions (IDRs) and lacks predictable functional motifs (39, 40). The Mediator complex forms liquid-liquid phase separation condensates at super-enhancers. These IDR-mediated condensates can be disrupted by 1,6-hexanediol (1,6-HD) (41, 42), an aliphatic alcohol, via inhibition of weak hydrophobic protein-protein interactions (43–47).

To determine whether individual Mediator subunits exhibit differential sensitivity to hydrophobic perturbation, we immunoprecipitated endogenous Mediator complexes from 293T cell lysates using antibodies against CDK8, MED1, and MED12 respectively, followed by washing with buffers containing increasing concentrations of 1,6-HD. Most of the detected subunits show modest decreases, whereas MED16 exhibits the most pronounced reduction in response to increasing 1,6-HD treatment (Fig 1A). To validate this observation, we performed parallel experiments using HeLa nuclear extracts (NEs) with 2,5-hexanediol (2,5-HD) as a negative control, as 2,5-HD does not disrupt weak hydrophobic interactions like 1,6-HD does. Consistently, MED16 dissociation was specifically induced by 1,6-HD but not by 2,5-HD in HeLa nuclear extract (Fig 1B).

Gel-filtration and 1,6-hexanediol treatment resolve a MED16-containing fraction that interacts with UBP1-TFCP2.

(A) Co-IP experiment with antibodies against endogenous CDK8, MED1, and MED12 in 293T whole cell lysis. The resultant immunoprecipitates were washed by washing buffers containing 0%, 5%, and 10% of 1,6-hexanediol respectively. Input: 0.5% of the total lysate was loaded. (B) Co-IP experiment with antibodies against endogenous CDK8, MED1, and MED12 in HeLa nuclear extract. The resultant immunoprecipitates were washed by washing buffers containing 10% of 1,6-hexanediol or 10% 2,5-hexanediol. Input: 0.2% of the total HeLa nuclear extract was loaded. (C) Co-IP experiment with antibodies against endogenous MED1 in 293T whole cell lysis. The resultant immunoprecipitated was washed by washing buffers containing 150 mM, 300 mM, and 500 mM of NaCl respectively. Input: 0.2% of the total HeLa nuclear extract was loaded. (D) Gel filtration chromatography of HeLa nuclear extract. 500µl HeLa nuclear extract was applied to Superose 6 column then was run in Buffer D. Column fractions of 500µl were collected. The Void (void volumn) of Superose 6 is based on the volume of effluent required for the elution of blue dextran (molecular mass of ∼2000 kDa). The molecular weight of corresponding fractions was detected by protein standards. (E) Immunoblots of anti-MED16 immunoprecipitation in HeLa nuclear extract and gel filtration factions No.29 to No.31. The immunoprecipitated proteins were detected with indicated antibodies by western blotting. HeLa NE input: 0.2% of the total HeLa nuclear extract was loaded. (F) Immunoblots of anti-TFCP2 immunoprecipitation in HeLa nuclear extract and gel filtration factions No.29 to No.31. The immunoprecipitated proteins were detected with indicated antibodies by western blotting. HeLa NE input: 0.2% of the total HeLa nuclear extract was loaded.

We next tested whether the MED16-Mediator interaction is mediated by ionic contacts by performing co-IP under increasing salt concentrations. Immunoprecipitated Mediator complexes were washed with buffers containing 150 mM, 300 mM, and 500 mM NaCl respectively. MED16 binding to the core complex remained unaltered under high-salt conditions (Fig 1C), indicating that MED16 integrates into the Mediator complex primarily through hydrophobic rather than electrostatic interactions. In parallel, we analyzed the elution profile of MED16 by gel-filtration chromatography of HeLa NE. Most Mediator subunits co-eluted at ∼2,000 kDa, corresponding to the intact holo-Mediator complex (Figure 1D). Notably, a substantial fraction of MED16 was also detected in lower molecular weight fractions (between 443 kDa and 150 kDa), well separated from the peak of core Mediator subunits (Figure 1D). Together, the 1,6-HD sensitivity and high-salt resistance indicate that MED16 associates with the Mediator complex primarily through hydrophobic interfaces. In addition, gel-filtration fractionation reveals that a portion of MED16 can be separated from the core Mediator complex in vitro.

UBP1 and TFCP2 associate with Mediator complex via MED16 subunit

MED16 has a molecular weight of approximately 95 kDa, but gel-filtration fractionation showed that the dissociated MED16 in fractions corresponds to 443-150 kDa (Fig 1D). This suggested that it may associate with other proteins. To identify these potential partners, we performed immunoprecipitation-coupled mass spectrometry (IP-MS) using antibodies against MED16, MED1, MED12, and CDK8 respectively. Two dimeric transcription factors, UBP1 and TFCP2, were identified as preferential MED16-binding proteins (Fig S1). Their co-elution with MED16 in the same gel-filtration fractions was confirmed by immunoblotting (Fig 1D). This suggests that UBP1 and TFCP2 may interact with the MED16 subunit (Fig S1).

To validate the association between UBP1/TFCP2 and MED16, an antibody specific to MED16 was used to capture the MED16 associated proteins in HeLa NE or in the 443-150 kDa fractions. In HeLa NE, MED1, MED23, TFCP2 and UBP1 were co-immunoprecipitated with anti-MED16 antibody, but only TFCP2 and UBP1 were co-immunoprecipitated with MED16 in 443-150 kDa fractions (Fig 1E). Reciprocally, anti-TFCP2 antibody is also able to pull down MED16 in either HeLa NE or 443-150 kDa fractions (Fig 1F). YY1 and HDAC1 are known to complex with TFCP2 and UBP1 to suppress HIV-1 transcription (34, 48, 49). YY1 was pulled down in both samples, whereas only a small amount of HDAC1 was pulled down from HeLa NE but not from the 443-150 kDa fractions (Fig 1E).

The above experiments established that UBP1 and TFCP2 interact with the biochemically resolved MED16. We next asked whether these transcription factors can also associate with the intact Mediator complex. To test this, we re-probed the IP samples from Figure 1B with antibodies against UBP1 and TFCP2. Both proteins were co-immunoprecipitated with the Mediator complex under 2,5-HD conditions, and this association was lost upon 1,6-HD treatment, which disrupts the MED16-Mediator interface (Fig 1B). These results indicate that UBP1 and TFCP2 can associate with the intact Mediator complex, likely through MED16.

To confirm that UBP1-TFCP2 associate with the Mediator complex specifically through MED16, we next mapped the functional domains responsible for MED16-UBP1/TFCP2 interactions. MED16 contains an N-terminal WD40-repeat (WDR) domain and a C-terminal αβ-domain. Serial truncation mutagenesis revealed that the N-terminal WDR domain is essential for MED16 integration into the core Mediator complex, whereas the C-terminal αβ-domain is strictly required for binding to UBP1-TFCP2 (Fig 2A, 2B).

The N-terminal WDR domain of MED16 anchors it within the Mediator, while its C-terminal αβ-domain binds UBP1.

(A) Schematic of Flag-MED16 deletion mutants. The WDR domains are shown as blue boxes. “1” represent full-length MED16 and “2” to “9” represent different truncations of MED16. All proteins were labeled with Flag-tag and expressed in 293T cell line as panel (B) indicated. (B) Domain mapping experiment of MED16. Flag-MED16 deletion mutants were coimmunoprecipitated with endogenous MED1 and TFCP2 using anti-Flag antibody. Asterisks indicate the IgG heavy and light chains. (C) Purified GST and GST-MED16 αβ-domain proteins were incubated with 6×His-tag TFCP2, UBP1, TFCP2L1, and YY1 at 40. After washing the interaction protein were eluted by boiling and immunoblotted with antibodies against GST-tag and His-tag. (D) Amino acids sequence analysis of UBP1 and TFCP2 protein, a 36 amino acids UBP1-specific sequence (USP) as shown below. (E) Schematic of Flag-UBP1 deletion mutants. FL represents full-length UBP1, △SAM represents SAM motif truncated UBP1, and DBD represents UBP1 DNA binding domain. (F) Domain mapping experiment of UBP1. Flag-UBP1 deletion mutants were coimmunoprecipitated with endogenous MED1, MED23, MED16, and TFCP2. Asterisks indicate the IgG heavy and light chains. (G) Schematic summary of the domain mapping results. The N-terminal WDR domain of MED16 mediates its integration into the Mediator complex, while the C-terminal αβ-domain binds UBP1. UBP1 interacts with MED16 through its UBP1-specific sequence (USP) and forms a heterodimer with TFCP2 via its SAM motif.

GST pull-down assays using purified recombinant proteins further validated direct binding. The GST-tagged MED16 αβ-domain exhibited robust direct interaction with His-tagged UBP1, with minimal binding to TFCP2, TFCP2L1, or YY1 (Fig 2C). Given the high sequence similarity between UBP1 and TFCP2, we further identified a unique 36-amino-acid specific peptide (USP) in UBP1 located between the DNA-binding domain and SAM motif (Fig 2D). Domain-mapping assays confirmed that the USP sequence is indispensable for MED16 binding, whereas the SAM motif mediates UBP1-TFCP2 dimerization (Fig 2E, 2F).

Together, these results establish that MED16 bridges UBP1/TFCP2 dimer to the Mediator complex. Specifically, the MED16 N-terminal WDR domain is responsible for anchoring MED16 to the Mediator, while MED16 C-terminal αβ-domain binds UBP1, with the USP sequence serving as the key recognition region for this interaction.

MED16 acts through the Mediator complex to support UBP1-activated transcription

Because the interaction between transcription factors and Mediator subunits is critical for target gene activation, we analyzed if Mediator MED16 is required for UBP1-driven transcription activation using reporter gene assay (Fig 3A). In the co-transfection assay, the full-length UBP1 significantly promoted transcriptional activation. However, the DNA-binding domain of UBP1 (DBD, 1-274aa) alone was unable to activate the reporter. The truncated UBP1 mutant (ΔSAM, 1-369aa), which contains the MED16 binding motif but lacks the SAM motif for dimerization with TFCP2, exhibited about half the transcriptional activation when compared to the full-length UBP1 (FL) (Fig 3B).

MED16 acts through the Mediator complex to support UBP1-mediated transcriptional activation.

(A) Schematic of the UBP1 luciferase reporter construct. (B) Reporter activity in HeLa cells co-transfected with the reporter and 300 ng of plasmid expressing full-length UBP1 (FL), the SAM motif-deleted mutant (ΔSAM), or the DNA binding domain alone (DBD). (C) Reporter activity in control (sh-NC) and MED16-knockdown (sh-MED16) HeLa cells with or without UBP1 overexpression. (D) Knockdown efficiency of sh-MED16 confirmed by qPCR and Western blotting. (E) Reporter activity in wild-type and MED16 knockout (MED16 KO) mouse pancreatic cancer HT cells with or without UBP1 overexpression. (F) MED16 knockout efficiency confirmed by Western blotting in five independent frameshift clones. (G) Dose-dependent effect of wild-type MED16 or MED16Δαβ on UBP1 reporter activity. (H) Combinatorial effects of MED16, UBP1, and TFCP2 overexpression on UBP1 reporter activity. (I) Schematic model of UBP1-mediated transcriptional activation. (Top) UBP1 recruits the Mediator complex to the promoter through MED16 to activate transcription. (Middle) Overexpression of UBP1 enhances Mediator complex recruitment, leading to more active transcription. (Bottom) Overexpression of MED16 sequesters UBP1 away from the endogenous Mediator complex (squelching), thereby reducing Mediator recruitment and inhibiting transcription. Firefly luciferase activity was normalized to Renilla luciferase activity. Data are presented as means ± SD from n = 3 independent experiments. *: p < 0.05, **: p-value<0.01 and “ns”: no significant.

To further confirm that MED16 is required for the transcriptional activity of UBP1, a HeLa cell line with MED16 knockdown was generated using viral-mediated shRNA targeting MED16, along with a negative control cell line (sh-NC) (Fig 3D). The UBP1-responsive reporter showed similar baseline of transcriptional activity in the two cell lines. The reporter was activated about 4-fold by overexpressing UBP1 in the negative control cell line, whereas knockdown of MED16 attenuated the activation to approximately two-fold (Fig 3C).

The coding sequences of human MED16 and Mouse Med16 share about 90% identity. Thus, we further established a Med16 knockout cell line generated using CRISPR-Cas9 gene editing method in a mouse pancreatic cancer cell line called HT. Five Med16 knockout clones (ko16a, ko16b, ko16c, ko16d, & ko16e), with different frameshift mutations were selected and verified by sequencing and Western-blotting (Fig 3F). All Med16-KO clones displayed 2-7-fold reduced reporter activity under ectopic UBP1 activation when compared to the wildtype HT cells, and Med16 KO exhibited decreased basal level of UBP1-reporter activity without UBP1 overexpression (Fig 3E and 3F). These results confirmed the specific requirement of MED16 for UBP1-mediated transcriptional activation.

To determine whether UBP1 requires the intact Mediator complex to activate transcription, we performed squelching experiments, based on the principle that overexpression of a Mediator subunit can sequester a transcription factor away from interacting with the endogenous Mediator complex, thereby inhibiting transcription 50-53). Wild-type MED16 repressed UBP1 reporter activity in a dose-dependent manner, consistent with this squelching effect. In contrast, increasing dose of MED16Δαβ, which cannot bind UBP1, had no inhibitory effect (Fig 3G). The same pattern was observed when UBP1 was co-expressed with TFCP2 (Fig 3H). These results indicate that UBP1-mediated transcriptional activation requires the intact Mediator complex, not MED16 acting alone.

Analysis of the UBP1 and MED16 co-regulated transcriptome

To characterize MED16-UBP1-regulated transcription at the genomic scale, we performed RNA-seq analyses in control and MED16-knockdown HeLa cells, combining with or without UBP1 overexpression. A total of 217 genes were significantly activated by UBP1 overexpression, and 60 genes were repressed (fold change > 1.5, p < 0.05). Among the UBP1-activated 217 genes, 198 were profoundly downregulated upon MED16 depletion, supporting that Mediator MED16 is largely required for UBP1-dependent transcriptional activation (Fig 4A-4C).

UBP1 and Mediator MED16 co-regulated transcriptome analysis.

(A) The number of genes repressed or activated by UBP1 overexpression in the sh-NC and shMED16 HeLa cell line (FC > |1.5|, p ≤ 0.05). (B) Venn diagram of UBP1-activated genes in the sh-NC HeLa cell line and the shMED16 HeLa cell line (FC > |1.5|, p ≤ 0.05). 198 genes which were activated by UBP1 overexpression but abolished by sh-MED16 were defined as UBP1 and MED16 co-regulated genes. (C) Heatmap showing the RNA-seq analysis of 217 UBP1-activated genes in sh-MED16 and sh-NC cell line treated with or without UBP1 overexpression. UBP1-activated genes were defined in sh-NC cell line under UBP1 overexpression at fold change>|1.5|, p-value≤0.05. (D) GO analysis of 198 UBP1 and MED16 co-regulated genes. The top 10 biological process items were presented. (E) Volcano map analysis of differentially expressed genes in sh-NC HeLa cell line with or without UBP1 overexpression. Genes failed to be activated by UBP1 in the sh-MED16 HeLa cell line we’re indicated by “×”. (F) UBP1 motif occurrence analysis of SFTPA1, SFTPA2, SFTPB and SFTPC. Motif occurrence were showed within the promoter region of the genes (−1000 bp to 200 bp relative to the TSS).

To explore the potential biological processes that are co-regulated by MED16 and UBP1, 198 co-regulated genes by MED16 and UBP1 were subjected to gene ontology (GO) analysis (Fig 4D and 4E). Interestingly, GO analysis revealed that the genes involved in the respiratory gaseous exchange process were regulated by MED16 and UBP1, such as SFTPA1, SFTPA2, SFTPB, and SFTPC (Fig 4F). These genes encode pulmonary surfactant associated proteins that promote alveolar stability by lowering the surface tension at the air-liquid interface in the peripheral air spaces. They are normally only expressed in alveolar epithelial cells and silenced in HeLa cells. Interestingly, highly matched UBP1 motifs were identified in their promoters (Fig 4F), and their activation by UBP1 overexpression is associated with the presence of these motifs. In our analysis, angiogenesis is also enriched in GO analysis (Fig 4D), which is consistent with the previous report that UBP1 plays a critical role in the regulation of extraembryonic angiogenesis (29, 54). Collectively, these results demonstrated that UBP1 and MED16 may collaborate to activate a target array of endogenous genes expression.

Mediator MED16 collaborates with UBP1 to inhibit HIV-1 transcription

Beyond their roles in transcriptional activation, UBP1 and TFCP2 have been reported to inhibit HIV-1 transcription through multiple mechanisms (32–34). We therefore asked whether MED16 participates in UBP1-mediated HIV-1 repression. To test this, we constructed an HIV-1 reporter driven by the viral core promoter (−78 to +60) upstream of luciferase (Fig 5A). Overexpression of MED16, UBP1 or YY1 reduced reporter activity, suggesting that MED16 behaves similarly to the known HIV-1 transcriptional repressors (Fig 5B). Overexpression of MED16 also inhibited the reporter in a dose-dependent manner (Fig 5E), whereas MED23 and MED24 had no effect (Fig 5C). Conversely, shRNA-mediated knockdown of MED16 modestly increased HIV-1 reporter activity (Fig 5D). These results indicate a specific role for MED16 in repressing HIV-1 transcription.

Mediator MED16 cooperates with UBP1 to inhibit HIV-1 reporter expression

(A) Schematic illustration of the HIV-1 reporter, with the HIV-1 core promoter and +1 to +60 TAR RNA sequence cloned upstream of the luciferase coding sequence (CDS). (B) The HIV-1 reporter was co-transfected with 1.5 μg of MED16, UBP1, TFCP2, YY1, or Flag-tag control plasmids in 293T cells. (C and D) Increasing amounts (0.7 μg, 1 μg, 1.5 μg) of MED16 or αβ-domain-deleted MED16 were transfected with the HIV-1 reporter. (E) The HIV-1 reporter was co-transfected with 1.5 μg of MED23, MED24, or Flag-tag control plasmids as additional controls. (F) Comparison of HIV-1 reporter luciferase activity between sh-NC and sh-MED16 293T cell lines. Firefly luciferase activity was normalized to Renilla luciferase activity. The luciferase expression levels between groups were compared using two-tail unpaired Student’s t-test, The data are presented as the means ± SDs from n = 3 independent experiment. * p < 0.05, ** p-value<0.01 and “ns”: no significant.

We next asked whether this repression requires the Mediator complex and the interaction with UBP1. The MED16ΔWDRs mutants, which cannot be integrated into the Mediator, failed to inhibit the HIV-1 reporter, indicating that MED16 must act through the Mediator complex to repress HIV-1 transcription (Fig 5F). A MED16 mutant lacking the C-terminal αβ-domain (MED16Δαβ), which cannot bind UBP1, also failed to inhibit the reporter (Fig 5G). Conversely, overexpression of the MED16 αβ-domain peptide, which competes with endogenous MED16 for UBP1 binding, de-repressed HIV-1 reporter activity (Fig H). Together, these results demonstrate that the functionality of MED16 in HIV-1 repression depends on both its integration into the Mediator complex and its interaction with UBP1.

UBP1 binding position determines the MED16-dependent inhibition of HIV-1 transcription

In a previous study, UBP1 was shown to inhibit HIV-1 transcription by recruiting YY1 and HDAC1(34). To investigate if MED16 employs a similar mechanism to repress HIV-1 transcription, we transfected HeLa cells with an HIV-1 reporter and treated them with panobinostat, a pan-HDAC inhibitor. Western blot analysis showed increased acetylation of histone H3 (H3ac) upon panobinostat treatment (Fig S2A). However, this treatment did not reverse MED16-mediated inhibition of HIV-driven luciferase activity (Fig S2A). This result suggests that MED16 may not rely on HDAC-mediated histone deacetylation to inhibit HIV-1 transcription and alternative mechanisms may be involved in HIV-1 transcription repression.

We then asked whether the cis regulatory elements in the HIV-1 promoter are necessary for MED16-mediated inhibition. Firstly, we examined whether the transactivating response region (TAR) plays a role in MED16-mediated HIV-1 inhibition HIV-1 inhibition. TAR is transcribed into TAR RNA, which recruits the host P-TEFb-containing super elongation complex (SEC) (55–57) to mediate the activation of viral protein Tat (58). UBP1 has been reported to restrict HIV-1 transcription at the level of elongation (33). A TAR RNA mutation HIV-1 reporter was created by replacing TAR loop nucleotides 31-34 UGGG with CAAA, which reduces binding affinity of Tat:P-TEFb and abolishes the function of TAR RNA (59). We observed that MED16 inhibited both wild-type and TAR-mutant reporter’s transcription, while mutated TAR attenuated UBP1-mediated inhibition (Fig S2B and S2C). This result suggested that MED16-mediated inhibition of HIV-1 is independent of the TAR-mediated HIV-1 transcriptional elongation.

HIV-1 proximal promoter has a string of GC-box that is bound by SP1, and an AP4 binding site between TATA-box and TSS (60). We then examined whether these two transcription factors are involved in the MED16-mediated transcription repression. A reporter driven by 4×GC-box was co-transfected with the MED16 expression plasmid, but no significant inhibition was observed compared with co-transfecting with the control vector (Fig S2D), suggesting no cross-talk between MED16 and Sp1 (GC-box). Then we mutated the AP-4 binding motif CAGCTG to CAGTCG (Fig S2E) to abolish AP-4 binding (61). Overexpressing MED16 inhibited the luciferase activity, regardless AP-4 binding status, suggesting that MED16 doesn’t work with AP-4 to inhibit the HIV-1 transcription.

Lastly, we examined if the UBP1-MED16 inhibition of HIV-1 transcription is due to the UBP1 binding position at the transcription started site (TSS) (32, 33). Two chimera reporters, both driven by a chicken β-ACTIN promoter, were created with distinctive UBP1 binding site position. In the first reporter (Fig 6A), the native TSS of the chicken β-ACTIN promoter was replaced with the HIV-1 TSS, which contains a UBP1 binding site. In the second reporter (Fig 6B), a UBP1 binding site was placed upstream of the chicken β-ACTIN TSS. Various truncated mutants of MED16 and UBP1 were co-transfected with these reporters, respectively. In the case of the first chimera reporter, transcription was inhibited by full-length MED16 and MED16-bound UBP1 variants (ΔSAM and FL-UBP1), but not by UBP1-binding deficient MED16 (MED16Δαβ) or MED16-binding deficient UBP1 (UBP1-DBD) (Fig 6A). Conversely, transcription of the second chimera reporter was not inhibited by any MED16 or UBP1 variants (Fig 6B). Additionally, comparison of transcription activities between the two reporters revealed that the first chimera exhibited approximately 50% lower transcription activity than the second, possibly due to the endogenous UBP1 activity (Fig 6C). These findings suggest that the UBP1 binding at the TSS seems to be key factor for MED16-mediated transcriptional repression

The position of the UBP1 binding site relative to the TSS dictates whether MED16-UBP1 activates or represses transcription.

Schematic showing a chimera reporter (A) driven by the Chicken β-ACTIN promoter, followed by an HIV-1 UBP1 binding site that also serves as the HIV-1 transcription start site and another chimera reporter (B) driven by the Chicken β-ACTIN promoter, a UBP1 binding site modify from HIV-1 TSS placed upstream of it. 1 μg of different truncation of MED16 or UBP1 plasmid was co-transfected with the reporter. The Luciferase activity comparison between two chimera reporters were shown in (C). “cβa-UBP1 bd” indicate the reporter that panel A described, and “UBP1 bd-cβa” indicate the reporter that panel B described. Firefly luciferase activity was normalized to Renilla luciferase activity. The luciferase expression levels between groups were compared using two-tail unpaired Student’s t-test, The data are presented as the means ± SDs from n = 3 independent experiment. * p < 0.05, ** p-value<0.01 and “ns”: no significant. UBP1 motif enrichment analysis of top 50 upregulated or top 50 downregulated transcripts induced by UBP1 were shown in (D). The transcripts changes levels are shown on the left as a heatmap, and average profile plots of UBP1 motif density are shown on the right. UBP1 motif occurrence analysis of UBP1 downregulated gene KIFC1 and GTF2H4 were shown in (E). Motif occurrence were showed within the promoter region of the genes (−500 bp to 200 bp relative to the TSS).

We then looked into whether the binding positional effect of UBP1 also impacted on the UBP1-tageted gene expression across the human genome. The correlation between the relative binding position of UBP1 to TSS and gene transcription levels was analyzed using the gene profiling data (Fig 4). We selected the top 50 upregulated and top 50 downregulated transcripts under UBP1 overexpression and analyzed the enrichment of the UBP1 motif around their promoters (Fig 6D). Notably, UBP1-activated transcripts showed significant enrichment of the UBP1 motif upstream and downstream of the TSS, while UBP1-repressed transcripts exhibited motif enrichment specifically overlapping the TSS (Fig 6D). For example, two UBP1-inhibited gene: KIFC1 and GTF2H4, were found to have their TSS overlapping with UBP1 binding motif (Fig 4F and 6E). Together with the chimeric reporter experiments, these data suggest that UBP1 can act as either a transcriptional activator or repressor, with its binding position relative to the TSS as a determinant of this two-way regulatory outcome.

UBP1-MED16 interrupts HIV-1 transcription initiation and sustains HIV-1 latency

We wondered how UBP1-MED16 binding within the proximal promoter might influence the HIV-1 transcription. To directly investigate how MED16-UBP1 binding at the TSS represses HIV-1 transcription, we performed in vitro immobilized template transcription assays using nuclear extracts from wild-type, Med16-knockout, and Med16-overexpressing HT cells (Fig 7A). During pre-initiation complex (PIC) assembly (without NTPs addition), MED16 overexpression reduced the recruitment of CDK8, MED12, MED1, and MED23 to the HIV-1 promoter. Upon transcription initiation by NTPs addition, Pol II Ser5 phosphorylation was markedly diminished, while the retention of TFIIB, a GTF that dissociates after initiation, was prolonged (Fig 7B). Functionally, MED16 overexpression reduced RNA synthesis of the viral template by approximately 45% (Fig 7C). These results suggest that MED16 overexpression disrupts the PIC formation for HIV-1 transcription initiation, leading to a reduction in RNA production.

MED16 interferes with HIV-1 PIC assembly at the TSS via its interaction with UBP1.

(A) Schematic of biotin-labeled immobilized template assay. (B) Immobilized template assay in wild-type, Med16 knockout and Med16 overexpression HT cell line. The binding proteins in individual steps were collected for immunoblotting. Beads: the streptavidin beads incubated with NE as a negative control; PIC: the pre-initiation step of transcription; Trx.: transcription initiation from the PIC step by adding NTPs. (C) qPCR quantification of generated RNA from different NE. The mRNA expression levels were compared between groups using two-tail unpaired Student’s t-test, The values are presented as the means ± SDs from n = 3 independent experiment. ** p-value<0.01 and “ns”: no significant.

Because the MED16-UBP1 interaction inhibits HIV-1 transcription (Figure 5), we went on exploring whether it might also play a role in maintaining HIV-1 latency in CD4+ T cell line. For this purpose, we used the J-Lat 10.6 cell line, a Jurkat-based cell line infected with a pseudotyped HIV-1 strain (HIV/R7/E-/GFP) (62). This cell line harbors the integrated HIV-1 copy in the second intron of the SEC16 gene (63) and the integrated HIV-1 copy remains silenced under normal conditions. HIV-1 latency reversal can be monitored by GFP expression that is controlled by viral promoter and serves as a marker for viral activation (Fig 8A). To induce viral expression, we treated cells with JQ1, a BET bromodomain inhibitor that reverses HIV-1 latency primarily by antagonizing BRD4-L-mediated suppression of the Tat-P-TEFb/SEC interaction (64, 65), while displacement of the BRD4-S isoform from the viral promoter may further contribute to this effect (66, 67).

Mediator MED16 prohibited PIC formation of HIV-1 transcription

(A) Schematic of the J-Lat 10.6 cell line, a Jurkat-derived CD4+ T cell model harboring an integrated latent HIV-1 provirus (HIV/R7/E-/GFP) in the second intron of the SEC16 gene. Under normal conditions, the provirus remains silenced, and latency reversal can be monitored by GFP expression driven by the viral promoter. (B) Western blot analysis confirming MED16 overexpression in MED16 OE J-Lat 10.6 cells compared with the control cell line. (C) Representative fluorescence microscopy images of MED16 OE and Control J-Lat 10.6 cells treated with or without 5 μM JQ1 for 48 hours. (D) Percentage of GFP-positive cells quantified by FACS. (E) Mean fluorescence intensity (MFI) measured using the BioTek Synergy 2 system. (F) Schematic model of MED16-mediated transcriptional regulation. MED16 uses its N-terminal WDR domain to anchor within the Mediator complex and its C-terminal αβ-domain to bind UBP1. When the UBP1 binding motif is located upstream of the TSS, UBP1 recruits the Mediator complex through MED16 to activate transcription. When the motif overlaps the TSS, as in the HIV-1 promoter, this interaction instead disrupts PIC assembly, leading to transcriptional repression and the maintenance of HIV-1 latency. The values are presented as the means ± SDs from n = 3 independent experiment. ** p-value<0.01

We generated J-Lat 10.6 cell lines with retroviral MED16 expression and a control line infected with the MSCV (Murine Stem Cell Virus) retrovirus (Fig 8B). Prior to JQ1 treatment, both cell lines showed minimal GFP expression, confirming that MED16 manipulation alone did not alter HIV-1 latency (Fig 8C). After treatment with 5μM JQ1 for 48h, the HIV-LTR-driven GFP expression is observed in both cell lines, but is reduced in the MED16 OE cell line, with a 40% decrease in the percentage of GFP-positive cells percentage and 0.54-fold reduction in MFI (mean fluorescence intensity) compared to the control cell line (Fig 8C, 8D, and 8E). These findings suggest that MED16 not only inhibits HIV-1 transcription but also plays a important role in sustaining HIV-1 latency in CD4+ T cells.

Discussion

In this study we identified that the UBP1-TFCP2 heterodimer interacts with Mediator subunit MED16. This interaction activates an array of endogenous gene expression and inhibits HIV-1 transcription.

Using gel-filtration chromatography and IP-MS, we first observed that transcription factor UBP1 and TFCP2 are associated with MED16 from the biochemically resolved fractions (Fig 1D, 1E and 1F). Although these transcription factors were initially found in MED16-containing fractions outside the core Mediator, our functional experiments demonstrate that they rely on the Mediator complex for transcriptional regulation. In the context of gene activation, squelching experiments showed that overexpressing wild-type MED16 inhibited UBP1-driven reporter activity, whereas the MED16△WDR and MED16△αβ mutants, which cannot either incorporate into Mediator or bind to UBP1, had no inhibitory effect (Fig 3G, 3H, 5G). This indicates that UBP1-mediated activation requires the intact Mediator complex, not MED16 alone. In the context of gene repression, MED16△WDR, a MED16 mutant lacking the N-terminal WDR domain, which cannot integrate into the Mediator, failed to inhibit the HIV-1 reporter (Fig 5F), demonstrating that UBP1-mediated repression also requires the whole Mediator complex. Thus, in either activating cellular genes or repressing HIV-1 transcription, UBP1 function by recruiting the Mediator complex through MED16 (Fig 8F). In this model, MED16 acts as a bridging subunit whose N-terminal WDR domain integrates it into the Mediator and whose C-terminal αβ-domain directly binds UBP1, thereby physically linking the transcription factor to the coactivator complex.

Previous studies have demonstrated UBP1’s role in inhibiting HIV-1 transcription through various mechanisms, including epigenetic and elongational regulation. Our investigation reveals a novel mechanism where the Mediator MED16-UBP1 interaction inhibits HIV-1 transcription initiation by compromising PIC formation. This finding highlights the importance of UBP1 binding site positioning, a concept recently underscored by Duttke and colleagues, who demonstrated that the spatial configuration of TF binding sites dictates transcriptional outcomes (68). Our work also extends this principle: UBP1 activates transcription when binding upstream or downstream of the TSS, but represses it when binding directly at the TSS. Additionally, we identified TFCP2 and the known HIV-1 silencing factor YY1 (34) in the MED16-containing fractions, although their precise roles in our mechanistic model remain to be determined.

Targeted inhibition of HIV-1 transcription, particularly through enhancing the silencing activity of UBP1-MED16 at the viral transcription start site (TSS), presents a potential tool for HIV-1 therapy. Many current approaches focus on inhibiting key regulatory proteins crucial for HIV-1 transcription (69–71). By strengthening the UBP1-MED16 interaction or increasing MED16’s affinity for the HIV-1 TSS, it may be possible to achieve more effective transcriptional silencing, which could reduce viral expression over prolonged periods. In the future, small molecule drug screening can be conducted targeting the enhancement of this interaction, and the obtained small molecules could be used in combination with existing antiretroviral therapies to enhance therapeutic efficacy (72, 73).

Interestingly, a small peptide fragment (αβ-domain) of MED16 has the opposing effect of activating HIV-1 transcription while potentially blocking the endogenous UBP1-MED16 interaction (Fig 5H). By manipulating the expressed form of MED16, we may be able to create a multi-faceted therapeutic tool that enables flexible regulation of HIV-1 transcription based on specific therapeutic needs. For instance, expressing the full-length MED16 could enhance the efficacy of antiretroviral therapies (ART), while expressing the MED16 αβ-domain peptide could support “shock-and-kill” strategies (74) by reactivating latent HIV-1, thereby enabling immune clearance or targeted antiretroviral intervention in previously hidden reservoirs. Together, these dual capabilities of MED16 may provide multiple strategies in the HIV-1 intervention.

Method

Plasmids

The human MED16, YY1, TFCP2, TFCP2L1 cDNAs were obtained from BRICS (Bio-Research Innovation center SUZHOU). The human UBP1 and mouse Med16 cDNAs were cloned from HeLa cells and HT cells respectively. Then full-length cDNA and truncations were cloned into 3×Flag-CMV-10 (Sigma-Aldrich-Aldrich) and pET-28b plasmids. The HIV-1 core promoter sequence were synthesized directly and clone into pGL3-Basic plasmid. All mutations on plasmid were generated by using FASR Site-Directed Mutagenesis Kit (Tiangen; Beijing; China). siMED16 and Ctrl shRNA oligonucleotides were cloned into pSiren-RetroQ (Clonetech).

Cell culture

HeLa and 293T cells were maintained in Dulbeccòs modified Eaglès medium (DMEM) containing 10% (v/v) FBS and 1% P/S. The HT cell line used in this study was an immortalized pancreatic cancer cell line derived from a KRASG12D P53+/-mouse and was generous provided by professor Xiaofei Yu in Fudan University. J-Lat 10.6 cell line was generous provided by professor Huanzhang Zhu in Fudan University. HT and J-Lat 10.6 cell were maintained in RPMI-1640 medium containing 10% (v/v) FBS and 1% P/S. All cells were maintained in the incubator with 5% CO2 at 37°C.

Western blot and real-time PCR assays

Method for Western blot and real-time PCR assay have been described previously (75). The primers for real-time PCR are listed in Table S1. Antibody for Western blot include the following MED23 (Abcam, ab200351), MED1 (Bethyl Lab, A300-793A), MED24 (Bethyl, A301-472A), MED6 (Santa Cruz, sc-9434), MED12 (Bethyl Lab, A300-774A), MED16 (Bethyl Lab, A303-668A), CDK8 (Abcam, ab115155), UBP1 (Protientech, 67318-1-Ig), TFCP2 (Proteintech, 15203-1-AP), YY1 (Proteintech, 66281-1-Ig), P300 (Abcam, ab14984), HDAC1 (Santa Cruz, sc-7872), PolII (Abcam, ab816), PolII-pSer5 (Millipore, 04-1572), TFIIB (Santa Cruz, sc-274D), TFIIH (Santa Cruz, sc-293), H3 (Abcam, ab1791), H3ac (Abcam, ab4729), β-Actin (Proteintech, 66009-1-Ig), GAPDH (Proteintech, 60004-1-Ig), Flag-tag (Sigma-Aldrich, F1804), His-tag (Proteintech, 66005-1-Ig), GST-tag (Invitrogen, 10004D).

Nuclear extract Preparation and gel-filtration chromatography assay

Preparation of un-dialyzed HeLa cell nuclear extract was described previously (76). Gel-filtration molecular weight markers (Sigma-Aldrich, MWGF1000) were loaded onto a Superose 6 10/300GL column to determine the void volume (Blue dextran ∼2000 kDa) and elution volume of protein standards (Thyoglobulin 669 kDa, Apoferritin 443kDa, Alcohol Dehydrogenase 150 kDa, Albumin 66 kDa). HeLa nuclear extract was applied to the Superose 6 10/300GL column and then was run in Buffer D (20 mM HEPES (pH 7.9), 300 mM KCl, 10% glycerol, 0.2 mM EDTA, 10 mM β-mercaptoethanol, and 1 mM PMSF). Fractions of 500µl were collected from the column and precipitated using trichloroacetic acid (TCA). Every third fraction was used for Western blot assay.

Co-immunoprecipitation (Co-IP) assay

For transient co-transfection, 293T cells plated in 10 cm dishes (90% confluency) were transfected with 10 μg of each plasmid using Lipofectamine 2000 (Invitrogen) according to the manufacturer’s protocol. After 48 h, the cells were harvested and lysed in lysis buffer (20mM HEPES pH7.5, 150mM NaCl, 1mM EDTA, 2.5mM EGTA, 0.3% Tween-20) with protease inhibitors. Lysates were subjected to immunoprecipitation with anti-Flag M2 beads (Sigma-Aldrich, A2220) and incubated at 4°C overnight, following by washing in lysis buffer three times. Then 50 μl 1× SDS loading buffer was added to the beads and boiled at 990 for 10 min for Western blot analysis with the indicated antibodies.

For endogenous Co-IP, 293T cells were seeded in 10 cm dishes and cultured under appropriate conditions. At nearly 80% confluence, the cells were harvested and lysed in lysis buffer with protease inhibitors. Lysate was centrifuged at 4°C, 13200 rpm for 10 minutes, and the supernatant was incubated with 4μg antibodies to MED16 (Bethyl, A303-668A), CDK8 (CST, 17396s), MED1 (Bethyl, A300-793A), MED12 (Bethyl, A300-774A) overnight. Then, 20 μl of Dynabeads Protein G (Invitrogen) was added and incubated for 2 hours at 4°C. The beads were then washed with lysis buffer (with 0%, 5%, 10% 1,6-hexanediol) three times and boiled with SDS loading buffer for Western blot assay with indicated antibodies.

As for the Co-IP experiment in HeLa nuclear extracts (NE) or gel-filtration fractions, the HeLa NE was first diluted with an equal volume of D300 buffer containing 20 mM HEPES (pH 7.9), 300 mM KCl, 10% glycerol, 0.2 mM EDTA, 10 mM β-mercaptoethanol, and 1 mM PMSF. Next, 4 μg of antibody was added directly to the diluted HeLa NE or gel-filtration fractions, followed by incubation at 4°C overnight. After the overnight incubation, 20 μL of Dynabeads Protein G (Invitrogen) was added and incubated for an additional 2 hours at 4°C. The beads were then washed three times with D300 buffer and boiled in SDS loading buffer for Western blot analysis with the specified antibodies.

IP-MS

After immunoprecipitation, the protein samples were separated by SDS0PAGE, running at 80 V for about 15 minutes. The electrophoresis was halted once the bromophenol blue dye front had traveled 0.5 cm into the resolving gel. The gel was then stained with Coomassie Brilliant Blue R250 to visualize protein bands, which were subsequently excised and diced into small fragments (1 mm³). These gel fragments underwent sequential washes: twice with a 50% acetonitrile (ACN) solution in 50 mM ammonium bicarbonate (NH4HCO3), followed by three washes with 10 mM NH4HCO3 and one wash with pure ACN to remove salts and detergents.

After dehydration with 100% ACN, the gel pieces were treated with 10 mM dithiothreitol (DTT) for 1 hour at 37 °C to reduce disulfide bonds. The DTT solution was then removed, and the gel fragments were dehydrated again with ACN. Next, 50 mM iodoacetamide (IAM) was added to alkylate free thiol groups, with incubation for 30 minutes in the dark. The gel pieces were washed three times with 10 mM NH4HCO3 and once with ACN before final dehydration.

For enzymatic digestion, the gel fragments were incubated overnight at 37 °C with trypsin (3 ng/μl in 10 mM NH4HCO3). The resulting peptides were extracted using a gradient ACN elution and analyzed by liquid chromatography-tandem mass spectrometry (LC-MS/MS) on an Orbitrap Fusion™ Lumos™ system.

IP-MS data analysis

IP-MS was performed using antibodies against MED16, MED1, MED12, and CDK8, each with three biological replicates. For each bait, protein abundance was normalized to the total intensity of all identified Mediator subunits in the same sample. This normalized abundance was expressed as parts per million (PPM), correcting for differences in immunoprecipitation efficiency across baits.

For each protein g in sample s, we first calculated the area-based abundance in parts per million: AreaPPMg,s= Area g,s∑g Area g,s×10 This value was then log2-transformed: Ag,s=log2⁡( AreaPPM g,s+1) To correct for differences in immunoprecipitation efficiency across samples, we computed a loading normalization factor Ls for each sample s, defined as the median log2 PPM of all identified Mediator subunits M in that sample: Ls=medianm∈M(Am,s) The normalized abundance of each protein was then obtained by subtracting this Mediator-derived normalization factor: Rg,s=Ag,s−Ls Finally, the bait-preferential score for each protein was defined as the difference between its normalized abundance in the MED16 IP and the mean of its normalized abundances in the other three Mediator bait IPs (MED1, MED12, CDK8): Score=Rg,MED16−mean(Rg,MED1,Rg,MED12,Rg,CDK8) A positive score indicates enrichment in the MED16 IP relative to the other three Mediator baits. Candidate proteins of interest were further filtered by the following criteria: detection in at least two of three replicates of the target bait; maximum unique peptide count ≥ 2; detection in at least one additional Mediator bait; exclusion of Mediator/Pol II subunits, ribosomal proteins, histones, keratins, and common contaminants; and annotation with transcription or DNA-binding functions using GO terms from org.Hs.eg.db and GO.db, requiring either transcription regulator activity or DNA binding combined with a transcription-related biological process.

Generation of Med16 KO cell line using CRISPR-Cas9

To generate Med16 knockout cell lines, the pX330-mCherry plasmid, containing the CRISPR-Cas9 machinery and target-specific guide RNAs, was used. A guide RNA sequence as reported in a previous study(77) was cloned into the pX330-mCherry vector. Cells were transfected with the pX330-mCherry plasmid using Lipofectamine 2000 (Invitrogen) according to the manufacturer’s protocol. After 48 hours, single mCherry-positive cells were sorted by fluorescence-activated cell sorting (FACS) into individual well of 96-wells plate. The sorted cells were then cultured and expanded. Individual clones were screened for gene knockout by sequencing and Western blotting to confirm successful editing.

Retrovirus infection

Stable cell lines with retroviral-mediated knockdown or overexpression were generated following the manufacturer’s protocol (Clontech) as described previously (13). Specific siRNA target sequences were designed using Thermo Scientific’s siRNA selection tools. Retroviruses were produced by co-transfecting 293T cells with pSiren-RetroQ (for knockdown) or pMSCV-puro (for overexpression) vectors along with the pCL10A1 helper plasmid, using Lipofectamine 2000 (Invitrogen). After 48 hours, the culture medium containing retroviruses was collected, supplemented with 20 μg/ml polybrene (Sigma-Aldrich-Aldrich), and filtered through a 0.45 μm filter. The filtered retroviruses were then added to target cells, which were subjected to spin infection at 2,500 rpm for 1.5 hours at 300. Twenty-four hours after infection, cells were selected with puromycin in 293T at 1 μg/ml and HeLa cells at 2.5 μg/ml.

Luciferase reporter assay

293T cells or HeLa cells were plated in a well of 12-wells plate (90% confluency) and transfected with 150 ng of luciferase reporter plasmid along with 50 ng pRL-TK (Promega) using Lipofectamine 2000 (Invitrogen). At 36-48h post transfection, the cells were harvested and subjected to dual-luciferase reporter assays according to the manufacturer’s protocol (Promega)

RNA-seq and data analysis

Total RNA was extracted from cells lysed with TRIZOL. We used a commercial RNA library kit to prepare the RNA-seq library according to the manufacturer’s specifications (VAHTS® Universal V8 RNA-seq LibraryPrep Kit) from Illumina. Briefly, mRNA was captured with mRNA capture beads before fragmentation. The RNA fragments were reverse transcribed and subjected to library amplification for Illumina (Novaseq 6000) sequencing.

For RNA-seq data analysis, the filtered clean reads were mapped to the human reference genome hg38 with HISAT2. Read counts for each gene were then quantified using HTSeq-count. Differential expression analysis was performed in R using DESeq2, which normalized the read counts and identified significantly differentially expressed genes.

Motif analysis

Motif enrichment analysis was performed using the closest known homolog of UBP1: the Tcfcp2l1 (GSE11431) position weight matrix (PWM) from the HOMER motif database (v4.9.1). Sequences and the Tcfcp2l1 PWM were submitted to the FIMO algorithm (MEME Suite v5.5.8) to scan for enriched motifs. This analysis generated base pair-resolution motif enrichment scores and associated p-values across all queried sequences.

In vitro transcription assays with immobilized template

The immobilized template assay was carried out based on the previous study(78). Biotin-labeled PCR products were generated from the HIV-1 reporter plasmid with biotin attached to the upstream end. These templates contain the HIV-1 core promoter region (−78 to +60) followed by the luciferase CDS. 6 μg of each biotinylated template was bound to 300 μg of Dynabeads M-280 Streptavidin (Dynal) according to the manufacturer’s instructions using a DynalMag™-2 (Dynal) for assistance. The immobilized templates were then resuspended in 100 μL of blocking buffer (20 mM HEPES, pH 7.9, 0.1 M KCl, 10% glycerol, 6 mM MgCl2, 2.5 mM DTT, 50 mg/mL BSA). After a 15-minute incubation at room temperature, 100 μL of nuclear extract from either WT, MED16-overexpressing, or Med16 KO HT cells was added and incubated at 30°C for 30 minutes to allow formation of pre-initiation complex (PIC). The immobilized templates were washed three times with buffer D containing 0.1% Triton X-100. The proteins bound to the templates were eluted with SDS gel loading buffer and analyzed by 10% SDS-PAGE.

For in vitro transcription, an NTP mix (25 mM each) was added to stimulate transcription after PIC formation, and the reaction was incubated at 30°C for 1 hour. The supernatant, containing the transcribed RNA products, was collected for qPCR quantification. The immobilized templates were again washed three times with buffer D plus 0.1% Triton X-100, and the proteins bound to the templates were eluted with SDS gel loading buffer and subjected to 10% SDS-PAGE.

Quantification and statistical analysis

For qPCR and luciferase assays, the means and SEMs were calculated from at least three independent experiments. Statistical significance was assessed using GraphPad Prism software, with data presented in graphs as means ± SDs. Student’s t-test was used for comparisons between two groups, with significance levels defined as *p < 0.05 and **p < 0.01.